---
title: "`r config$publication_top_name`"
pagetitle: NHS Help with Health Costs
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---
  
  <style type="text/css">
  
  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
</style>
  <html lang="en">
  
```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

<div class = "quartoheader", id = "fixed-header">
        <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
    </div>

<main>
  
# Background Information and Methodology {.toc-ignore}
  
Published `r config$publication_date`
  
  
------------------------------------------------------------------------
  
## 1. Background information
The NHSBSA administer a range of services to help people with the health costs, include help with:

* NHS prescription costs
* NHS dental costs
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Full details of these schemes are available on the [NHS Help with Health Costs webpage](https://www.nhsbsa.nhs.uk/nhs-help-health-costs).

Six different certificate schemes are available, with different levels of help with health costs:

* [Maternity exemption certificates](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/maternity-exemption-certificates)
* [Medical exemption certificates](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/medical-exemption-certificates)
* [NHS Tax Credit Exemption Certificates](https://www.nhsbsa.nhs.uk/exemption-certificates/nhs-tax-credit-exemption-certificates)
* [NHS Low Income Scheme](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme)
* [NHS Prescription Prepayment certificate     (PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc)
* [NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT     PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)

These statistics provide estimates of volumes related to these certificate schemes. Data covers the financial years 2019/2020 to 2023/24, where the financial year runs from April 1 to March 31. The time period covered by these statistics includes the Coronavirus (COVID-19) pandemic period. It is worth noting that during this time period there was a reduction in face to face primary care consultations, referrals, new prescriptions and diagnoses due to the COVID-19 pandemic and national lockdowns.  

#### Users and uses
This publication can have a wide range of uses, including:

* informing government or local NHS policy
* monitoring the uptake of NHS Help with Health Costs schemes
* allowing scrutiny by the public and stakeholders

The NHS Business Services Authority (NHSBSA) administer services providing help with health costs. The data that forms the basis of these statistics is collected as a by-product of this process.

## 2. Methodology

### 2.1 Applicant and certificate holder details {.toc-ignore}

#### 2.1.1 Age
Reported ages represent the age of the applicant when they submitted the application.

Some certificates may also provide support for partners and dependent children. In these cases, the age will represent the age of the lead applicant only.

It may not be possible to calculate an age if the applicant's date of birth was not supplied or was incorrect. If the captured date of birth looks incorrect, the age will classified as "Not Available". Age restrictions will differ between services, as defined in the relevant "methodology" sections.

Due to the method used to calculate age using the applicants date of birth, some data may change between releases. Historic age data may change if a person updates their data to correct inaccuracies with their date of birth.

#### 2.1.2  Location (Country/Integrated Care Board)

These statistics use the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about) to identify Integrated Care Board (ICB) and country for each individual. This uses the captured postcode information supplied for the application/certificate.

Postcodes are typically provided during the application process. Some applicant's may then update postcode details when addresses change. The postcode used for these statistics will represent the latest known postcode information.

Some NHS Help with Health Costs services are generally limited to applicants from England whilst other services may also provide certificates for Scotland, Wales and Northern Ireland. Within the summary report, figures will represent all activity. Where applicable the supporting dataset will split the country by either England, Other or Unknown. 

ICB reporting is limited to applicants that have an identifiable postcode in England.

Due to the method used to assign location, some data may change between releases. Historic location data may change:

* if a person moves postcode and updates their details between releases
* if the NSPL reclassifies a postcode against a different area in the new NSPL edition

#### 2.1.3  Deprivation
Deprivation reporting uses the [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019). In particular, the headline Index of Multiple Deprivation (IMD). IMD data combines with the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about) using lower-layer super output area (LSOA) 2011. This identifies the general deprivation for an applicant's postcode.

As deprivation is based on indices calculated for England only, reporting is limited to applicants that can be identified with an English postcode.

The measure of deprivation reported in these statistics is the IMD quintile. Each quintile represents 20% of the population, based on groups of LSOA. If an applicant's postcode does not map to the NSPL the data will be reported as 'unknown' IMD quintile.

Postcodes are usually provided during application process. Some applicant's may then update postcode details when addresses change. The postcode used for these statistics will represent the latest known postcode information.

Due to the method used to assign IMD quintile, some data may change between releases. Historic location data may change: 

* if a person moves postcode and updates their details between releases to a LSOA in a different IMD quintile
* if the NSPL reclassifies a postcode against a different quintile in the new NSPL edition

### 2.2 Application and certificate count details {.toc-ignore}

#### 2.2.1 Application counts
Where a count of applications is reported, this includes all applications received regardless of application route, status or outcome.

A single applicant could submit multiple applications during the same period. Each application will be treated as a unique event and counted accordingly.

For the NHS Low Income Scheme and NHS Tax Credit Exemption Certificates, a single application could cover multiple people. However, the figures in this report represent certificate level data and therefore figures will represent a single application, regardless of how many individuals were included on the application.

#### 2.2.2 Issued certificate counts
Where a count of issued certificates has been provided this will only include cases where the application has been completed and a certificate issued to the applicant.

These counts are based solely on certificates issued by NHSBSA. There is no ability to identify if the certificate was received and then used by the applicant.

Some certificates may be reissued if the certificate holder changes there personal information or requests a replacement due to damage or loss of a certificate. These additional (reissued) certificates will not be included in any certificate counts. However, if a certificate holder submits a completely new application, and receives an additional certificate, this will be counted alongside all other issued certificates.

For the NHS Low Income Scheme and NHS Tax Credit Exemption Certificates, a single certificate could represent multiple people. However, the figures in this report represent certificate level data and therefore figures will represent a certificate, regardless of how many individuals are covered.

### 2.3 Service specific data capture methodology {.toc-ignore}

#### 2.3.1 NHS Low Income Scheme
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_lis,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.1.1 Date restrictions and exclusions applied
The number of applications and issued outcomes cover the period 2018/19 to 2023/24.

For analysis based on demographics such as age and location, data relates to the latest financial year of 2023/24.

The following restrictions apply to all reporting relating to NHS Low Income Scheme:

* Limited to HC1 applications submitted to NHS Low Income Scheme.
* Excluding any HC5 (refund only) applications.
* Excluding any data relating to certificates issued to asylum seekers assisted by UK Visas and Immigration (UKVI). For these individuals a HC2 is automatically issued as part of the support they receive.

##### 2.3.1.2 Certificates issued
Applications are only considered as complete if the data shows an outcome has been issued to the applicant. Applications that are currently in progress are not included. Applications abandoned before completion are only included in figures relating to applications received.

The certificate type for issued certificates will be HC2 or HC3.

##### 2.3.1.3 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands, with applicants aged 65 and over grouped as 65+.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For NHS Low Income Scheme data this includes where the calculated age would be less than 15 or greater than 99.

##### 2.3.1.4 Certificate duration
Duration is the number of months between the start and expiry date for a certificate. To simplify reporting, age bands reported represent the most common scenarios.

##### 2.3.1.5 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based on mid-year population estimates published by ONS for 2022, limited to population aged 16 and above. 

#### 2.3.2 Maternity exemption certificate
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_hes,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.2.1 Date restrictions and exclusions applied
The number of applications and issued outcomes cover the period 2018/19 to 2023/24.

For analysis based on demographics such as age and location, data relates to the latest financial year of 2023/24.

Applications that are still ongoing or were abandoned before completion are only included in figures relating to applications received.

##### 2.3.2.2 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands, with applicants aged 45 and over grouped as 45+.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For maternity exemption data this includes where the calculated age would be less than 15 or greater than 60.

##### 2.3.2.3 Certificate duration
Maternity exemption certificates are valid until a year after the birth of the child. The overall duration of the certificate will depend on how early the application is submitted. Certificates issued very early during pregnancy could be valid for up to a total of 22 months. Certificates only issued after the birth would be valid for less than 12 months.

Duration is calculated as the number of months, rounded to the nearest month, between the certificate start and expiration date. If the calculated duration would suggest potential data issues, figures will be reported as "Not Available".

When applications are received during pregnancy the expiration date will be calculated as one year after the expected due date. If a baby is born late, the certificate holder can apply for an extension to have the expiration date revised to reflect the actual date of birth. 

Due to the method used to calculate certificate duration, some data may change between releases. Historic data may change: 

* if an applicant applies for an extension when a baby is born late to extend the certificate duration to 12 months after the birth

##### 2.3.2.4 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based on mid-year population estimates published by ONS for 2022, limited to female population aged 16 to 45.

#### 2.3.3 Medical exemption certificate
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_hes,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.3.1 Date restrictions and exclusions applied
The number of applications and issued outcomes cover the period 2018/19 to 2023/24.

For analysis based on demographics such as age and location, data relates to the latest financial year of 2023/24.

Applications that are still ongoing or were abandoned before completion are only included in figures relating to applications received.

##### 2.3.3.2 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands. The medical exemption certificate provides support with NHS prescription charges and therefore is only relevant between the ages of 16 and 59 when prescription items are not automatically free based on patient age.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For medical exemption data this includes where the calculated age would be less than 15 or greater than 60.

##### 2.3.3.3 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based estimated patient counts, aged 16-59, receiving any NHS prescription items during the reported time period.

#### 2.3.4 NHS Prescription Prepayment Certificate (PPC)
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_hes,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.4.1 Date restrictions and exclusions applied
The number of applications and issued outcomes cover the period 2018/19 to 2023/24.

For analysis based on demographics such as age and location, data relates to the latest financial year of 2023/24.

Applications that are still ongoing or were abandoned before completion are only included in figures relating to applications received.

##### 2.3.4.2 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands. The PPC provides support with NHS prescription charges and therefore is only relevant between the ages of 16 and 59 when prescription items are not automatically free based on patient age.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For PPC data this includes where the calculated age would be less than 15 or greater than 60.

##### 2.3.4.3 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based estimated patient counts, aged 16-59, receiving any NHS prescription items during the reported time period.

#### 2.3.5 NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC)
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_hrt,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.5.1 Date restrictions and exclusions applied
The HRT PPC was introduced in April 2023. All analysis in this publication reflects activity during 2023/24.

Applications that are still ongoing or were abandoned before completion are only included in figures relating to applications received.

##### 2.3.5.2 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands. The HRT PPC provides support with NHS prescription charges and therefore is only relevant between the ages of 16 and 59 when prescription items are not automatically free based on patient age.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For HRT PPC data this includes where the calculated age would be less than 15 or greater than 60.

##### 2.3.5.3 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based estimated patient counts, aged 16-59, receiving any NHS prescription items for drugs that were eligible for support by the HRT PPC during the reported time period.

#### 2.3.6 NHS Tax Credit Exemption Certificates
Reporting reflects data available in the NHSBSA Data Warehouse environment as of `r format(as.Date(config$extract_date_hes,"%Y%m%d"),"%d %B %Y")`.

##### 2.3.6.1 Date restrictions and exclusions applied
The number of issued certificates cover the period 2018/19 to 2023/24.

For analysis based on demographics such as age and location, data relates to the latest financial year of 2023/24.

##### 2.3.6.2 Age
The age reported will represent the age of the applicant at the time of application. Figures are reported in 5 year age bands, with applicants aged 65 and over grouped as 65+.

An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data. For NHS Low Income Scheme data this includes where the calculated age would be less than 15 or greater than 90.

##### 2.3.6.3 ICB base population
For ICB reporting, rates have been calculated per 10,000 base population. This base population is based on mid-year population estimates published by ONS for 2022, limited to population aged 16 and above.

### 2.4 Base population identification {.toc-ignore}

When reporting activity by geographic area, such as ICB, figures can be standardised to adjust for area size. This is done by calculating a rate using an appropriate population baseline. Different population baselines have been applied, depending on the most appropriate population for the service area.

#### 2.4.1 NHS prescribing data patient counts

Medical exemption and prescription prepayment certificates (PPC and HRT PPC) are focused on providing support with NHS prescription costs. For these services the base population is based on estimated patient counts based on data captured from NHS prescribing data. 

The patient count estimates are based on available data from the NHS prescription dataset maintained by NHSBSA. Prescription data is collected from the submission of prescriptions by dispensing contractors to the NHSBSA. These prescriptions can be issued by GPs and other authorised prescribers such as nurses, dentists, and allied health professionals. Prescriptions that are issued by hospitals can also be dispensed in the community and submitted for reimbursement. Prescriptions that are issued in hospitals and fulfilled by the hospital pharmacy or dispensary are not included in this data.

The estimated patient counts included in this publication may omit patients whose prescribing activity either cannot be identified or cannot be attributed to a geographic area or age group. For 2023/24, the required patient information could be identified for 92% of prescription items dispensed in England.

##### 2.4.1.1 Geographic area identification

For prescribing via the Electronic Prescription Service (EPS), the patient's residential address is captured from the prescription data. This address is then assigned to a lower-layer super output area (LSOA) 2011. The LSOA can then be used to identify both the ICB and the deprivation profile (see section 2.1.3) based on mapping available in the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about).

LSOA is only captured for prescriptions via EPS. For patients receiving both paper and EPS prescribing, the LSOA data is based on their EPS activity only.

Some patients may be captured with data against multiple residential addresses. To prevent patient's being counted multiple times, in these instances patients have been assigned to a single location. This is based on data captured against the latest available prescription record in the time period.

##### 2.4.1.2 Patient age

To report a patient’s age consistently across a financial year these statistics classify a patient based on their age on the 30th of September of that given financial year. This uses date of birth information from Personal Demographic Service (PDS) for verified NHS numbers only. 

Information relating to NHS numbers that have not been verified by PDS were excluded from patient counts as these patients could not be accurately assigned to an age group. 

In rare circumstances patient data may be updated resulting in multiple dates of birth being captured for an NHS number. In these instances, the date of birth captured against the latest available prescription record in the time period was used to calculate the patient's age.

Where certificates are providing support with NHS prescription costs, these certificates may only be relevant between the ages of 16 and 59 where patients are not automatically exempt from prescription charges based on age. Therefore, estimated patient counts are restricted to patients between the ages of 16 and 59.

##### 2.4.1.3 Drug classification (HRT PPC population only)

The NHSBSA holds drug and prescription data at a pack level and uses this information for the correct reimbursement of dispensing contractors. The drug database maintained by NHSBSA includes flags to identify which drugs are licensed to treat menopause in the UK and covered by the HRT PPC.

For population baseline figures used for HRT PPC, prescribing activity is restricted to only count patients who received any of the medication covered by the HRT PPC. Although medication could be added or removed from the qualifying medication list, medicines were included in the patient count analysis if they were covered by the HRT PPC at any point during the reported time period.

For population baseline figures for certificates other than HRT PPC, no drug restrictions have been applied.

#### 2.4.2 ONS published population estimates

Maternity exemption certificates, NHS Tax Credit Exemption Certificates and the NHS Low Income Scheme provide support with a range of health costs. For these services the base population is based on population counts as published by ONS.

The population estimates were sourced from [population estimates for health geographies dataset published by ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates).

These population estimates are based on mid-year estimates, with 2022 being the latest available year at time of reporting. Figures published by ONS include a breakdown by age and gender. Data has been aggregated to include the most appropriate combination of age and gender for the service being reported. For NHS Tax Credit Exemption Certificates and the NHS Low Income Scheme population figures are based on people aged 16 and over. For maternity exemption certificates the population figures are based on the female population aged 15-45.

#### 2.4.3 ONS live birth statistics

To provide context for the maternity exemption certificate some live birth statistics have been reported. These figures were sourced from data published by ONS. It is not suitable, or possible, to use these figures as an accurate representation of the number of potential claims for maternity exemption certificates. Therefore this figures should only be considered as additional context.

There are some considerations to recognise when comparing live births statistics with maternity exemption certificate data:

* Live births will not include all pregnancies which could be eligible for a maternity exemption certificate.
* The time periods can not be accurately aligned. ONS publish live birth statistics by calendar year, with 2021 being the latest available year (at time of reporting) for the breakdowns required in this publication. Additionally, maternity exemption certificates cover a period longer than a single year as they cover both the pregnancy and upto one year following birth.

The distribution of live births by age of mother was sourced from [ONS nomis tool query for live births by characteristics of mother](https://www.nomisweb.co.uk/query/construct/summary.asp?mode=construct&version=0&dataset=203).

The distribution of live births by IMD decile was sourced from [birth characteristics dataset published by ONS (see worksheet 12)](https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/birthcharacteristicsinenglandandwales/2021/birthcharacteristics2021workbooknew.xlsx).

## 3. Strengths and limitations

### 3.1. Strengths {.toc-ignore}

The main strength of these statistics is the completeness of the dataset and accuracy of information captured during processing activities carried out by the NHSBSA. This dataset covers all applications, and subsequent outcomes, for support with health costs via certificate schemes administered by NHSBSA.

All the data has come from the same administrative source. Using one source of data ensures that the data is complete and all data required for this publication is present.

### 3.2. Limitations {.toc-ignore}

These statistics are reliant on the quality of data supplied and processed during the application process. 

Certificates issued for the NHS Low Income Scheme and NHS tax credit exemption may also provide coverage to partners and dependent children. However, any demographic analysis will be limited to using the information for the lead applicant or certificate holder.

Due to the lack of a consistent, reliable unique person identifier, figures cannot be published relating to the number of individuals applying to, or receiving support from, any of the NHS Help with Health Costs service areas. Additionally, as an individual could apply for, and receive, multiple certificates in a single time period the figures included in these statistics may include multiple records relating to the same individual.

In addition to the lack of a consistent unique person identifier, information governance restrictions prevent the linking of data across different services and therefore it is not possible to provide any data relating to the number of individuals receiving support across all NHS Help with Health Costs services. It is also possible that an individual could hold multiple certificates from different services at the same point in time.

## 4. Revisions

Any revisions that we make to these statistics will be made in line with our [Revisions and Corrections policy](https://www.nhsbsa.nhs.uk/policies-and-procedures). Any significant errors that are identified within these statistics after their publication that would result in the contradiction of conclusions previously drawn from the data will be notified of prominently on our website and any other platforms that host these statistics, corrected as soon as possible, and communicated clearly to users and stakeholders.

## 5. Related statistics, comparability and useful resources

### 5.1 Code of Practice for Statistics {.toc-ignore}

These statistics have been produced in compliance of the Code of Practice for Statistics. You can find more on the code of practice and its pillars, principles and practices from the [UK Statistics Authority website](https://code.statisticsauthority.gov.uk/).

## 6. Quality of the statistics

We aim to provide users of this publication with an evidence-based assessment of its quality and the quality of the data from which it is produced. We do so to demonstrate our commitment to comply with the UK Statistics Authority’s (UKSA) Code of Practice for Statistics, particularly the pillar of Quality and its principles.

**Q1 Suitable data sources** – Statistics should be based on the most appropriate data to meet intended uses. The impact of any data limitations for use should be assessed, minimised, and explained.

**Q2 Sound methods** – Producers of statistics and data should use the best available methods and recognised standards and be open about their decisions.

**Q3 Assured quality** – Producers of statistics and data should explain clearly how they assure themselves that statistics and data are accurate, reliable, coherent, and timely.

This is an assessment of the quality of these statistics against the European standard for quality reporting and its dimensions specific to statistical outputs, particularly:
  
*	Relevance
*	Accuracy and reliability
*	Timeliness and punctuality
*	Accessibility
*	Coherence and comparability

### 6.1. Relevance {.toc-ignore}

***This dimension covers the degree to which the product meets user need in both coverage and content***
  
The HwHC publication summarises the number of certificates issued by NHSBSA to provide support to individuals with their health costs. We believe that this can be used to inform policy decisions at a national and local level, by the public to scrutinise uptake of NHS Help with Health Costs services, and by academia and applied health researchers for matters relating to public health. The NHSBSA also routinely receives Freedom of Information requests and parliamentary questions about this subject matter. 

We will be gathering feedback from users of these statistics on an on-going basis to help shape them and ensure that they remain relevant and useful.


### 6.2 Accuracy and Reliability {.toc-ignore}

***This dimension covers the statistics proximity between an estimate and the unknown true value***
  
#### 6.2.1. Accuracy  {.toc-ignore}
  
These statistics are derived from data collected during processing activities carried out by the NHSBSA to issue certificates to provide help with health costs. 

The figures used are collected as an essential part of the process of supply certificates to individuals.

As there is a manual data entry element to the services covered by this publication, inevitably some small errors may occur in the data. The NHSBSA take measures to minimise these error including monthly quality checks based on a random sample of certificates issued. 

As of May 2024 the 12 month rolling accuracy results for NHS Help with Health Costs service areas were:
```{r HwHC Accuracy}
data.frame(
  'HwHC Service' = c("Maternity exemption", "Medical exemption", "NHS Low Income Scheme", "PPC", "HRT PPC"),
  '12 month rolling accurary (%)' = c("98.4","98.5","98.3","98.2","96.3"),
  check.names=FALSE
) |> 
  knitr::kable(
    align = "lr"
  )
```

#### 6.2.2. Reliability {.toc-ignore}

Data can be collected and updated throughout the full lifecycle, from initial application to certificate expiration. These statistics are derived from the latest available information at the time the report was developed.


### 6.3. Timeliness and punctuality {.toc-ignore}

***Timeliness refers to the time gap between publication and the reference period. Punctuality refers to the gap between planned and actual publication dates***
  
The publication date is determined by the availability of the data, dependent on the completion of processing activities.

A decision was made to allow a window of a couple of months following the end of the financial year to accommodate applications submitted by third parties, for example, pharmacy PPC sales.


### 6.4. Accessibility and clarity {.toc-ignore}

***Accessibility is the ease with which users can access the data, also reflecting the format in which the data are available and the availability of supporting information. Clarity refers to the quality and sufficiency of the metadata, illustrations, and accompanying advice***
  
The statistical summary narrative for this publication is presented as an HTML webpage, with supporting documentation also released in HTML format. Summary data and additional analysis is presented in tables in Excel files.

A supporting online dashboard has also been created to allow some interrogation of the available data.

The R code used to produce the newest release of the HwHC publication is also available from the [NHSBSA GitHub](https://github.com/nhsbsa-data-analytics).

***Clarity***
  
A glossary of terms is included in this document.


### 6.5. Coherence and comparability {.toc-ignore}

***Coherence is the degree to which data have been derived from different sources or methods but refer to the same topic or similar. Comparability is the degree to which data can be compared over time and domain***
  
The HwHC publication is the only statistics available on the certificates issued by NHSBSA to provide support with health costs. The statistics are all derived from the same administrative data source with the same methodology applied to all data points.

#### 6.5.1. Comparisons over time {.toc-ignore}

In order to allow for comparisons to be made over time these statistics cover the whole period for which data is available, from financial year 2018/19 onwards.

Changes to the figures over time should be interpreted in the wider context of the national landscape. Uptake of these support services administered by NHSBSA may be impacted by a number of external forces including the different support schemes available and changes to the health and financial status of individuals.

***Trade-offs between output quality components
This dimension describes the extent to which different aspects of quality are balanced against each other***

The main trade-off in this publication is the balance between timeliness and data quality. Sufficient time is allowed from the data being made available to allow for the information to be produced and quality assured.

***Assessment of user needs and perceptions
This dimension covers the processes for finding out about users and uses and their views on the statistical products***

Alongside the release of these statistics the NHSBSA will also be releasing a continuous feedback survey, allowing users to quickly tell us their thoughts on the content and utility of these statistics. This feedback, along with feedback gathered from other routes such as direct contact, will be used to shape the content and style of future HwHC publications and other statistical products from the NHSBSA.

***Performance, cost and respondent burden
This dimension describes the effectiveness, efficiency and economy of the statistical output***

There is no respondent burden for HwHC data, as all data are extracted from existing NHSBSA information and transactional systems. 
This initial release has been developed with a reproducible analytical pipeline (RAP) in mind and RAP principles applied where possible. This development has been done in R and the code used will be made publicly available at the [NHSBSA GitHub](https://github.com/nhsbsa-data-analytics) in due course. Further development is planned to the RAP used for this publication to automate as many tasks as possible. 

***Confidentiality, transparency and security
The procedures and policy used to ensure sound confidentiality, security and transparent practices***

Trustworthy statistics and the data behind them are an important part of well-informed decision making, and are vital to support improvement across the wider health and social care system. It is accepted, however, that where statistics provide information on small numbers of individuals, the NHS Business Services Authority have a duty under data protection law to avoid directly or indirectly revealing any personal details. In addition, NHSBSA staff members are required to adhere to relevant NHS data confidentiality guidelines.

The NHSBSA has robust confidentiality and security policies that were adhered to during the production of these statistics. More information on these policies and how we follow them can be found in our [Confidentiality and Access Statement](https://www.nhsbsa.nhs.uk/sites/default/files/2021-05/Confidentiality_and_Access_Statement_v002.pdf).

A risk assessment around potential disclosure of personal identifiable information through these statistics was carried out during their production. In line with the NHSBSA’s Statistical Disclosure Control Policy, information where a count of individuals less than five can be inferred, has been redacted.

## 7. Glossary of terms used in these statistics

**Age**

Age, and therefore age bands, have been calculated based on the applicant's date of birth and the date the application was received. An age band of 'Not Available' has been used where the applicants age could not be confidently assigned from the available data.

**IMD Quintile**

The [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) have been used to provide a measure of deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals' living conditions.

The reported IMD quintile, where 1 is the most deprived and 5 the least deprived, is derived from the postcode currently held for an applicant. There are a small number of records each year that we have reported as having an unknown IMD quintile. These are applicants with no fixed abode or with british forces addresses, or where we have been unable to match the applicant postcode to an English postcode in the [National Statistics Postcode Lookup (NSPL) - May 2023](https://geoportal.statistics.gov.uk/datasets/379c875939ec4fc4bb48ed32711eb042/about).

IMD quintiles are calculated by ranking census lower-layer super output areas (LSOAs) from most deprived to least deprived and dividing them into equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

**Integrated Care Board (ICB)**

Integrated care boards (ICBs) are a statutory NHS organisation responsible for developing a plan in collaboration with NHS trusts/foundation trusts and other system partners for meeting the health needs of the population, managing the NHS budget and arranging for the provision of health services in the defined area. They took over the functions of Clinical Commissioning Groups (CCG) in July 2022.

**Number of applications**

The number of applications includes all applications received, regardless of status or outcome. Where an application may cover multiple people, the application will only be counted once. 

Where an applicant submits multiple applications, each application will be counted as a distinct event and counted individually.

Some certificate attributes, such as certificate type and duration, may not be available at the point of application and therefore no application figures can be reported against these attributes.

**Number of issued certificates**

The number of issued certifications includes all cases where a certificate was issued to the customer, regardless of current status. It is not possible to confirm that the issued certificate was then used by recipient. Where a certificate may cover multiple people, the certificate will only be counted once.

Some individuals may receive multiple certificates, with each certificate being counted as an distinct event.

## 8. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous, and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:
  
**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**
  
NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

</main>
  
