---
title: "`r config$publication_top_name`"
pagetitle: Help with Health Costs
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

```{=html}
<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>
```
```{=html}
<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
```
<html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date`

## Key findings

T.B.C

------------------------------------------------------------------------

## 1. Introduction

### 1.1. Background {.toc-ignore}

The NHSBSA administer a range of services to help people with the health costs, include help with:

* NHS prescription costs
* NHS dental costs
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Full details of these schemes are available on the [NHS Help with Health Costs webpage](https://www.nhsbsa.nhs.uk/nhs-help-health-costs).

Six different certificate schemes are available, with different levels of help with health costs:

* [NHS Prescription Prepayment certificate     (PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc)
* [NHS Tax credit exemption certificate](https://www.nhsbsa.nhs.uk/exemption-certificates/nhs-tax-credit-exemption-certificates)
* [Medical exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/medical-exemption-certificates)
* [Maternity exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/maternity-exemption-certificates)
* [NHS Low Income Scheme (LIS)](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme)
* [NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT     PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)

These statistics provide estimates of volumes related to these certificate schemes. Data covers the financial years 2019/2020 to 2023/24, where the financial year runs from April 1 to March 31.



------------------------------------------------------------------------

## 2. Results and commentary

### 2.1 NHS Low Income Scheme

`r infoBox_border("",text = "The NHS Low Income Scheme (LIS) provides income related help. Support is available for people who do not receive help with health costs via an existing exemption. This is a means tested benefit, but also considers council tax and housing costs.

NHS Low Income Scheme exemption provides help with:

* NHS prescriptions
* NHS dental treatment
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

The amount of support provided is based on weekly income, necessary outgoings, plus any savings or investments. Issued certificates will also cover the applicant's partner and dependent children under 20 years old."
, width = "100%")`

#### Figure X: Estimated number of applications received 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 374,000 applications received in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Applications decreased 4% from 2022/23</b>", width = "100%")`
:::
:::

```{r ch_lis_fy_app}
ch_lis_fy_app
```

`r get_download_button(title = "Download chart data", data = dl_lis_fy_app, filename = "hwhc_figure_x")`

Applicants will either complete a paper HC1 form or apply via the [online application for LIS](https://apply-for-help-with-nhs-costs.nhsbsa.nhs.uk/apply-online).

Applications will include new applicants and existing certificate holders reapplying. The number of applications will also include incomplete applications.

Applications to the NHS Low Income Scheme decreased slightly in 2023/24, with 374,000 applications compared to 388,000 applications in 2022/23. The number of applications is notably higher than between 20219/20 and 2021/22.

In 2023/24, applications to the NHS Low Income Scheme decreased compared to the previous year. There were 374,000 applications in 2023/24, compared to 388,000 thousand applications in 2022/23. But the number of applications is still higher in 2023/24 than between 20219/20 and 2021/22.

#### Figure X: Estimated number of outcome decisions by support type 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>almost two-thirds of the 261,000 decisions issued in 2023/24 provided full help with health costs</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>16% of decisions issued in 2023/24 resulted in no support to the applicant</b>", width = "100%")`
:::
:::

```{r ch_lis_fy_issue}
ch_lis_fy_issue
```

`r get_download_button(title = "Download chart data", data = dl_lis_fy_issue, filename = "hwhc_figure_x")`

Two levels of support are available from the NHS Low Income Scheme. Depending on the outcome of the assessment, applicants could receive:

* an [HC2 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc2-certificates-full-help-health-costs) for full help with health costs, including free NHS prescriptions.
* an [HC3 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc3-certificates-limited-help-health-costs) for limited help with health costs. A HC3 certificate will show how much the holder has to pay towards health costs.

In some cases the outcome of the application assessment will not result in a HC2 or HC3 certificate. This includes people with existing means tested benefits that include health cost exemption. Also, applicants may not receive a certificate based on the financial assessment outcome. 

There were 261,000 decisions issued in 2023/24. This included 161,000 HC2 certificates and 58,000 HC3 certificates. Compared to 2022/23 there has been a slight decrease in the number of certificates issued.

#### Figure X: Estimated number of active NHS Low Income Scheme certificates 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 492,000 active NHS Low Income Scheme certificates in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>There were more active certificates in 2023/24 than 2022/23. This is despite a drop in the number of certificates issued.</b>", width = "100%")`
:::
:::

```{r ch_lis_fy_active}
ch_lis_fy_act
```

`r get_download_button(title = "Download chart data", data = dl_lis_fy_act, filename = "hwhc_figure_x")`

Active certificates measures the number of certificates valid in the specified financial year. Some NHS Low Income Scheme certificates can be valid up to 5 years. Therefore, the number of active certificates are higher than the certificates issued. When certificates expire a certificate holder could reapply. They may then end up with two (or more) active certificates for a single financial year.

Trends in active certificates are generally in line with certificates issued. In 2023/24 there were 355,000 HC2 certificates active and 137,000 active HC3 certificates.

#### Figure X: Estimated number of issued NHS Low Income Scheme certificates 2023/24, split by certificate type and duration {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over 80% of HC3 certificates were valid for 12 months</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>84% of all certificates are valid for either 6 or 12 months</b>", width = "100%")`
:::
:::

```{r ch_lis_duration}
ch_lis_duration
```

`r get_download_button(title = "Download chart data", data = dl_lis_duration, filename = "hwhc_figure_x")`

The duration of NHS Low Income Scheme certificates can vary from a little as a day to a maximum length of 5 years. 

The circumstances of the applicant will determine the duration of the certificate. This will reflect the chances of circumstances changing, with some common scenarios:

* Pensioners receiving full help may receive a 5 year certificate.
* Non final-year students may receive a 12 month certificate.
* Applicants in receipt of non-means tested benefit may receive a 12 month certificate.
* Applicants earning a wage may receive a 6 month certificate.
* Students nearing the end of their course may receive a short term certificate.

Almost 48,000 HC3 certificates, issued in 2023/24, were valid for a period of 12 months. This represents over 83% of all HC3 certificates issued in 2023/24. HC2 certificates show more variation in length the HC3 certificates. In 2023/24 43,000 HC2 certificates were valid for 6 months and 92,000 were valid for 12 months.


#### Figure X: Estimated number of issued NHS Low Income Scheme certificates 2023/24 split by certificate type and applicant age band {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over half of the certificates issued in 2023/24 were to people aged under 25 or 65 and over</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24 more certificates were issued to applicants aged 65 and over than any other age group</b>", width = "100%")`
:::
:::

```{r ch_lis_age}
ch_lis_age
```

`r get_download_button(title = "Download chart data", data = dl_lis_age, filename = "hwhc_figure_x")`

NHS Low Income Scheme certificates cover the applicant, partners and any dependent children. Age information is only captured for the lead applicant though. 

Pensioners are more likely than other age groups to receive a long term certificate. This will impact the age distribution of certificates issued in a year.

The younger and older age groups receive more certificates than other ages. Almost 52,000 certificates went to people aged 65 and over in 2023/24. A further 45,000 went to applicants 20-24 and 29,000 to applicants aged 19 and under.

#### Figure X: Estimated number of issued NHS Low Income Scheme certificates 2023/24, split by certificate type and IMD quintile {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24 60,000 certificates went to people in the most deprived areas</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24, the second quintile received almost as many HC2 certificates as the most deprived quintile</b>", width = "100%")`
:::
:::

```{r ch_lis_imd}
ch_lis_imd
```

`r get_download_button(title = "Download chart data", data = dl_lis_imd, filename = "hwhc_figure_x")`

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

Certificates were issued in the most deprived areas more than three times more often than the least deprived areas. 

Despite being means-tested, certificates are still issued to applicants from less deprived areas. In 2023/24, over 13,000 HC2 certificates and over 5,000 HC3 certificates went to the least deprived areas.

#### Figure X: Estimated number of issued NHS Low Income Scheme certificates per 10,000 population by ICB 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The three ICBs in north London had the highest rate of certificates issued per 10,000 population</b>", width = "100%")`
:::
:::

```{r vis_icb, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(map_lis_icb, ch_lis_icb, ncol = 2)
```

`r get_download_button(title = "Download chart data", data = dl_lis_icb, filename = "hwhc_figure_x")`

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report.

North West London ICB had the highest estimated number of certificates issued per 10,000 population. In 2023/24, an estimated 79 certificates were issued per 10,000 population in this ICB.

Surrey Heartlands ICB had the lowest estimated number of certificates issued per 10,000 population in 2023/24. This ICB had an estimated 25 certificates per 10,000 population.

------------------------------------------------------------------------

## 3 About these statistics

### 3.1 Statistical Disclosure Control (SDC) {.toc-ignore}

Statistical disclosure control has been applied to these statistics. Volume figures have been redacted in the supporting summary tables if they relate to fewer than 5 items. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

### 3.2. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases rounding has been to two significant figures in the key findings and information boxes and three significant figures in the main text. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 3.3. Index of deprivation {#about_imd .toc-ignore}

The [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) have been used to provide a measure of deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals' living conditions.

The reported IMD quintile, where 1 is the most deprived and 5 the least deprived, is derived from the postcode currently held for an applicant. There are a small number of records each year that we have reported as having an unknown IMD quintile. These are applicants with no fixed abode or with british forces addresses, or where we have been unable to match the applicant postcode to an English postcode in the [XXX citation needed](https://geoportal.statistics.gov.uk/datasets/national-statistics-postcode-lookup-may-2022/about).

IMD quintiles are calculated by ranking census lower-layer super output areas (LSOAs) from most deprived to least deprived and dividing them into equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

------------------------------------------------------------------------

## 4 Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:
**Email**:
[accessibility\@nhsbsa.nhs.uk](mailto:accessibility@nhsbsa.nhs.uk){.email}

**Phone**: 0191 203 5318

[Find out about call
charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We'll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the 'Contact us' section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

------------------------------------------------------------------------

## 5 Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](%60r%20config$stats_survey_link%60) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:
**Email:**
[statistics\@nhsbsa.nhs.uk](mailto:statistics@nhsbsa.nhs.uk){.email}

**You can also write to us at:**
NHSBSA - Statistics\
NHS Business Services Authority\
Stella House\
Goldcrest Way\
Newburn Riverside\
Newcastle upon Tyne\
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>
