---
title: "`r config$publication_top_name`"
pagetitle: Help with Health Costs
always_allow_html: true
output:
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    anchor_sections: false
    css: www/style.css
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{=html}
<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>
```
```{=html}
<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
```
<html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date`

## Key findings (2023/24)

### NHS Low Income Scheme {.toc-ignore}

* The number of applications and issued certificates in 2023/24 was a slight decrease from 2022/23 but higher than the two years prior. There were 161,000 HC2 certificates, and 58,000 HC3 certificates, issued in 2023/24.

* Certificates were most commonly issued to people from the most deprived areas of the country and people from the younger (15-24) and older (65+) age groups.

### Maternity exemption certificates {.toc-ignore}

* The number of applications and certificates issued has risen over the past few years. There were 552,000 maternity exemption certificates issued in 2023/24.

* Almost two-thirds of medical exemption certificates were issued to women by 7 months before the expected due date. Less than 5% of certificates were issued after the expected due date.

* The number of maternity exemption certificates issued in the most deprived areas is a little lower than may be expected based on the distribution of live births.

### Medical exemption certificates {.toc-ignore}

* The number of applications and certificates issued has risen over the past few years following a drop during 2019/20 and 2021/22. There were 475,000 medical exemption certificates issued in 2023/24.

* The is a strong relationship with age, with the number of certificates issued rising as age rises.

### NHS Prescription Prepayment Certificate (PPC) {.toc-ignore}

* The number of PPC certificates purchased has increased in each of the past 5 years. The 12-month certificate is more popular than the 3-month certificate, with 2,070,000 12-month certificates issued in 2023/24 compared to 936,000 3-month certificates issued.

* The is a strong relationship with age, with the number of certificates issued rising as age rises.

* In 2023/24 the five London ICBs were the bottom five in terms of certificates issued per 10,000 population.

### NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC) {.toc-ignore}

* The HRT PPC was launched in April 2023, with 131,000 certificates being issued in the first month. This represents 23% of the 566,000 certificates issued in 2023/24.

* There is a strong relationship with deprivation, with the number of certificates issued rising as deprivation decreases. This relationship is similar to the pattern seen for patient's receiving prescribing of drugs licensed to treat menopause in the UK which are covered by the HRT PPC.

* The South West region has some of the largest rates of HRT PPCs issued. In 2023/24, 5 of the 7 ICBs in the South West region were in the top 10 for number of HRT PPCs issued per 10,000 population.

### NHS tax credit exemption certificates {.toc-ignore}

* The number of NHS tax credit exemption certificates has been declining for a number of years. As part of the Universal Credit rollout people are being migrated away from tax credits and onto new benefit schemes. In 2023/24 1,150,000 certificates were issued, compared to 3,740,000 in 2018/19.

* The age profile of recipients has changed over the past few years. In 2018/19, 55% of certificates were issued to people aged under 40 years old but in 2023/24 this has dropped to 30% of issued certificates.

------------------------------------------------------------------------

## 1. Things you should know

### 1.1. Scope {.toc-ignore}

`r infoBox_border("Background",text = "The NHSBSA administer a range of services to help people with the health costs, include help with:

* NHS prescription costs
* NHS dental costs
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Full details of these schemes are available on the [NHS Help with Health Costs webpage](https://www.nhsbsa.nhs.uk/nhs-help-health-costs).

Six different certificate schemes are available, with different levels of help with health costs:

* [NHS Prescription Prepayment certificate     (PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc)
* [NHS Tax credit exemption certificate](https://www.nhsbsa.nhs.uk/exemption-certificates/nhs-tax-credit-exemption-certificates)
* [Medical exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/medical-exemption-certificates)
* [Maternity exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/maternity-exemption-certificates)
* [NHS Low Income Scheme (LIS)](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme)
* [NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT     PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)"
, width = "100%")`

These statistics provide estimates of volumes related to these certificate schemes. Data covers the financial years 2018/2019 to 2023/24, where the financial year runs from April 1 to March 31.

These statistics are based on information from applications and data captured during certificate processing activities. Therefore, where information has not been supplied or captured accurately, this may impact the quality of the statistics.

Figures has been reported split by individual help with health cost service. Although an individual could use multiple services, data governance rules prevent data from being linked across different service areas. 

Some of the help with health cost services covered in this publication are provided only for residents in England while some services are also provided for other countries in the United Kingdom. Within this summary report figures will represent all activity, with additional breakdowns by country available in the supporting datasets.

------------------------------------------------------------------------

## 2. Results and commentary

### 2.1 NHS Low Income Scheme

`r infoBox_border("",text = "The NHS Low Income Scheme (LIS) provides income related help. Support is available for people who do not receive help with health costs via an existing exemption. This is a means tested benefit, but also considers council tax and housing costs.

NHS Low Income Scheme exemption provides help with:

* NHS prescriptions
* NHS dental treatment
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

The amount of support provided is based on weekly income, necessary outgoings, plus any savings or investments. Issued certificates will also cover the applicant's partner and dependent children under 20 years old."
, width = "100%")`

#### Figure X: Number of applications received 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 374,000 applications received in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Applications decreased 4% from 2022/23</b>", width = "100%")`
:::
:::

```{r lis_application_chart}
lis_application_objs$chart
get_download_button(title = "Download chart data", data = lis_application_objs$chart_data, filename = "hwhc_figure_x")
```

Applicants will either complete a paper HC1 form or apply via the [online application for LIS](https://apply-for-help-with-nhs-costs.nhsbsa.nhs.uk/apply-online).

Applications will include new applicants and existing certificate holders reapplying. The number of applications will also include incomplete applications.

Applications to the NHS Low Income Scheme decreased slightly in 2023/24, with 374,000 applications compared to 388,000 applications in 2022/23. Despite this decrease, the number of applications received still higher than previous years.

Applications to the NHS Low Income Scheme are assessed to determine if the applicant qualifies for any support.

#### Figure X: Number of outcome decisions by support type 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>almost two-thirds of the 261,000 decisions issued in 2023/24 provided full help with health costs</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>16% of decisions issued in 2023/24 resulted in no support to the applicant</b>", width = "100%")`
:::
:::

```{r lis_issued_chart}
lis_issued_objs$chart
get_download_button(title = "Download chart data", data = lis_issued_objs$chart_data, filename = "hwhc_figure_x")
```

Two levels of support are available from the NHS Low Income Scheme. Depending on the outcome of the assessment, applicants could receive:

* an [HC2 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc2-certificates-full-help-health-costs) for full help with health costs, including free NHS prescriptions.
* an [HC3 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc3-certificates-limited-help-health-costs) for limited help with health costs. A HC3 certificate will show how much the holder has to pay towards health costs.

In some cases the outcome of the application assessment will not result in a HC2 or HC3 certificate. This includes people with existing means tested benefits that include health cost exemption. Also, applicants may not receive a certificate based on the financial assessment outcome. 

There were 261,000 decisions issued in 2023/24. This included 161,000 HC2 certificates and 58,000 HC3 certificates. Compared to 2022/23 there has been a slight decrease in the number of certificates issued.

The proportion of outcome decisions that do not include a HC2 or HC3 certificate has risen from 11% in 2018/19 to 16% in 2023/24.

An outcome decision is only issued to the applicant following the completion of the assessment. Therefore these figures will not include any applications that are ongoing or were abandoned before completion. 

#### Figure X: Estimated number of active NHS Low Income Scheme certificates 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 492,000 active NHS Low Income Scheme certificates in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>There were more active certificates in 2023/24 than 2022/23. This is despite a drop in the number of certificates issued.</b>", width = "100%")`
:::
:::

```{r lis_active_chart}
lis_active_objs$chart
get_download_button(title = "Download chart data", data = lis_active_objs$chart_data, filename = "hwhc_figure_x")
```

Active certificates measures the number of certificates valid in the specified financial year. Some NHS Low Income Scheme certificates can be valid up to 5 years. Therefore, the number of active certificates are higher than the certificates issued. 

When certificates expire a certificate holder could reapply, and may then end up with two (or more) active certificates within a single financial year.

The number of active certificates is only available from 2019/20 due to retention schedules for certificates prior to this period. It should be noted that active certificates are based solely on the start and expiration date of the issued certificates. This does not consider if the certificate holder was using the certificate or not.

Trends in active certificates are generally in line with certificates issued. In 2023/24 there were 355,000 HC2 certificates active and 137,000 active HC3 certificates.

#### Figure X: Number of NHS Low Income Scheme certificates issued, split by certificate type and duration (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over 83% of HC3 certificates were valid for 12 months</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>84% of all certificates are valid for either 6 or 12 months</b>", width = "100%")`
:::
:::

```{r lis_duration_chart}
lis_duration_objs$chart
get_download_button(title = "Download chart data", data = lis_duration_objs$chart_data, filename = "hwhc_figure_x")
```

The duration of NHS Low Income Scheme certificates can vary from a little as a day to a maximum length of 5 years. 

The circumstances of the applicant will determine the duration of the certificate. This will reflect the chances of circumstances changing, with some common scenarios:

* Pensioners receiving full help may receive a 5 year certificate.
* Non final-year students may receive a 12 month certificate.
* Applicants in receipt of non-means tested benefit may receive a 12 month certificate.
* Applicants earning a wage may receive a 6 month certificate.
* Students nearing the end of their course may receive a short term certificate until the end of their course.

Almost 48,000 HC3 certificates, issued in 2023/24, were valid for a period of 12 months. This represents over 83% of all HC3 certificates issued in 2023/24. HC2 certificates show more variation in length than HC3 certificates. In 2023/24 43,000 HC2 certificates were valid for 6 months and 92,000 were valid for 12 months.

#### Figure X: Number of NHS Low Income Scheme certificates issued, split by certificate type and applicant age band (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over half of the certificates issued in 2023/24 were to people aged under 25 or 65 and over</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24 more certificates were issued to applicants aged 65 and over than any other age group</b>", width = "100%")`
:::
:::

```{r lis_age_chart}
lis_age_objs$chart
get_download_button(title = "Download chart data", data = lis_age_objs$chart_data, filename = "hwhc_figure_x")
```

NHS Low Income Scheme certificates cover the applicant, partners and any dependent children. Age information is only captured for the lead applicant though. 

Pensioners are more likely than other age groups to receive a long term certificate. This will impact the age distribution of certificates issued in a year as pensioners may only have to apply once every 5 years.

The younger and older age groups receive more certificates than other ages. Almost 52,000 certificates went to people aged 65 and over in 2023/24. A further 45,000 went to applicants 20-24 and 29,000 to applicants aged 19 and under. For HC3 certificates in particular this is stronger relationship with age with only 24% of certificates issued to people aged 25 to 64.

For outcome decisions where no certificate was issued there is less variation by age and more activity for people in the 30 to 54 age group. In 2023/24, over a quarter of the decisions issued to people aged 30 to 54 had no certificate issued.

#### Figure X: Number of NHS Low Income Scheme certificates issued, split by certificate type and IMD quintile (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24 60,000 certificates went to people in the most deprived areas</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24, the second quintile received almost as many HC2 certificates as the most deprived quintile</b>", width = "100%")`
:::
:::

```{r lis_imd_chart}
lis_imd_objs$chart
get_download_button(title = "Download chart data", data = lis_imd_objs$chart_data, filename = "hwhc_figure_x")
```

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure. These indices of deprivation are specific to England and therefore can only be applied to certificates that can be assigned to an English postcode.

Certificates were issued in the most deprived areas more than three times more often than the least deprived areas. 

The distribution of outcome decision is similar within each IMD quintile. This suggests that deprivation profile alone is not a strong indicator of whether or not an application will result in a HC2, HC3 or no certificate being issued. However, it should be noted that this will not include applications that were abandoned before being completed.

Despite being means-tested, certificates are still issued to applicants from less deprived areas. In 2023/24, over 13,000 HC2 certificates and over 5,000 HC3 certificates went to the least deprived areas.


#### Figure X: Number of NHS Low Income Scheme certificates issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The three ICBs in north London had the highest rate of certificates issued per 10,000 population</b>", width = "100%")`
:::
:::

```{r lis_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  lis_icb_objs$map, 
  lis_icb_objs$chart, 
  ncol = 2
)
```
```{r lis_icb_vis_download}
get_download_button(title = "Download chart data", data = lis_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. These are the latest available population statistics at the time of publication. To align with the people who may benefit from the NHS Low Income Scheme, population figures are based on people aged 16 and over.

The map and chart are based on the total number of HC2 and HC3 certificates issued in 2023/24. This does not include any application outcomes where a HC2 or HC3 was not issued. Figures for individual application outcome types are included in the accompanying data.

North West London ICB had the highest estimated number of certificates issued per 10,000 population. In 2023/24, an estimated 79 certificates were issued per 10,000 population in this ICB. The top three ICBs in terms of certificates issued per 10,000 population were all from north London.

Surrey Heartlands ICB had the lowest estimated number of certificates issued per 10,000 population in 2023/24. This ICB had an estimated 25 certificates per 10,000 population.

------------------------------------------------------------------------

### 2.2 Maternity exemption certificate

`r infoBox_border("",text = "Maternity exemption certifications exempt the holders from NHS prescription charges. It can also be used as as proof of entitlement to free NHS dental treatment.

Maternity exemption certificates are available to anybody who is currently pregnant, or has given birth in the last 12 months. The certificate is valid until 12 months after the birth of the baby.

Applications are submitted via a midwife, doctor or health visitor who confirms eligibility. Both paper and digital applications are available."
, width = "100%")`

#### Figure X: Number of applications received 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 555,000 applications received in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Applications increased 11% from 2022/23</b>", width = "100%")`
:::
:::

```{r matex_application_chart}
mat_application_objs$chart
get_download_button(title = "Download chart data", data = mat_application_objs$chart_data, filename = "hwhc_figure_x")
```

Applications for maternity exemption certificates dipped in 2020/2021 but has increased in each of the following years. In 2023/24, the increase in applications of 11% is the largest in recent years.

In 2023/24 there were 555,000 applications received, up from 499,000 in 2022/23.

Applications are submitted via a healthcare professional, who confirms eligibility. Therefore, only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

#### Figure X: Number of certificates issued 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 551,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Issued certificates increased 12% from 2022/23</b>", width = "100%")`
:::
:::

```{r matex_issued_chart}
mat_issued_objs$chart
get_download_button(title = "Download chart data", data = mat_issued_objs$chart_data, filename = "hwhc_figure_x")
```

The number of issued maternity exemption certificates dipped in 2020/2021 but has increased in each of the following years. In 2023/24, the increase in certificates issued of 12% is the largest in recent years.

In 2023/24 there were 551,000 maternity exemption certificates issued, up from 494,000 in 2022/23.

Maternity exemption certificates are valid from when a pregnancy is confirmed up to one year following birth. Therefore, some certificates may be valid across multiple financial years.

#### Figure X: Estimated number of active maternity exemption certificates 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 1,260,000 maternity exemption certificates active in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The number of active certificates each financial year is gradually increasing</b>", width = "100%")`
:::
:::

```{r matex_active_chart}
mat_active_objs$chart
get_download_button(title = "Download chart data", data = mat_active_objs$chart_data, filename = "hwhc_figure_x")
```

The number of active certificates is only available from 2019/20 due to retention schedules for certificates prior to this period. It should be noted that active certificates are based solely on the start and expiration date of the issued certificates. This does not consider if the certificate holder was using the certificate or not.

There have been over 1,100,000 maternity certificates active in each of the past five financial years. The 1,260,000 certificates active in 2023/24 is a 4% increase on the 1,210,000 certificates active in 2022/23.

Some certificates may be active in multiple financial years, and the number of years a certificate will cover depends on how early on during pregnancy the certificate is issued.

#### Figure X: Proportion of certificates issued relative to expected due date (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over 60% of certificates were issued by 7 months before the expected due date</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Only 5% of certificates were issued after the expected due date.</b>", width = "100%")`
:::
:::

```{r ch_mat_duration}
ch_mat_duration
get_download_button(title = "Download chart data", data = dl_mat_duration, filename = "hwhc_figure_x")
```

Maternity exemption certificates can be issued from the date the pregnancy is confirmed up to one year after the birth. Applications are submitted by a healthcare professional on the applicants behalf. Any delays in this process will impact the length of support available with some women not receiving a certificate until later in the pregnancy or until after the birth.

Over 60% of certificates issued in 2023/24 had been issued 7 months prior to the expected delivery date. This period would align with the stage in pregnancy that women start to interact with NHS maternity services. 

In almost 95% of cases, the certificate had been issued by the expected due date. Although only a small proportion of certificates were issued after the due date, this still represents 29,500 certificates where the mother did not have the maternity exemption certificate until after giving birth.

#### Figure X: Number of maternity exemption certificates issued, split by age band of recipient (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Numbers peak in the 30-34 age band, with 198,000 certificates issued to this age group</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Distribution by age is similar to the pattern seen for live birth statistics</b>", width = "100%")`
:::
:::

```{r matex_age_chart, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  mat_age_objs$chart |> 
    highcharter::hc_title(text = 'Number of maternity exemption certificates issued (2023/24)'), 
  ch_births_age |> 
    highcharter::hc_title(text = 'Number of live births (2021)') |> 
    highcharter::hc_subtitle(text = 'Source: Office for National Statistics'), 
  ncol = 2
)
```
```{r matex_age_chart_download}
get_download_button(title = "Download chart data", data = mat_age_objs$chart_data, filename = "hwhc_figure_x")
```

The distribution of maternity exemption certificates by age peaks at 30-34, with 198,000 certificates issued to this age group in 2023/24. 

There is no available data that accurately shows the number of women eligible for maternity exemptions in a specific time period. However, the age distribution is very similar to latest available live birth statistics published by Office for National Statistics.

#### Figure X: Number of maternity exemption certificates issued, split by deprivation profile (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>More certificates were issued to people from the second most deprived quintile than any other</b>", width = "100%")`
:::
::: col-md-6
`r infoBox_no_border("", text = "<b>In the most deprived areas the number of certificates issued may be lower than expected compared to live birth figures alone</b>", width = "100%")`
:::
:::

```{r matex_imd_profile_chart, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  mat_imd_objs$chart |> 
    highcharter::hc_title(text = 'Number of maternity exemption certificates issued (2023/24)'), 
  ch_births_imd |> 
    highcharter::hc_title(text = 'Number of live births (2021)') |> 
    highcharter::hc_subtitle(text = 'Source: Office for National Statistics'), 
  ncol = 2
)
```
```{r matex_imd_profile_chart_download}
get_download_button(title = "Download chart data", data = mat_imd_objs$chart_data, filename = "hwhc_figure_x")
```

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

In general the number of maternity exemption certificates decreases in line with deprivation. The exception is the most deprived areas, where the 118,000 certificates issued is slightly less than 123,000 issued to the second most deprived group.

The latest available live birth statistics published by Office for National Statistics suggest that in general the number of births are higher in the more deprived areas. In the most deprived areas, women may be more likely to qualify for other income-related benefits that offer the same support as a maternity exemption certificate.

#### Figure X: Number of maternity exemption certificates issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>Figures range from 408 certificates per 10,000 population in Cheshire and Merseyside ICB to 636 certificates per 10,000 population in Bedfordshire, Luton and Milton Keynes ICB</b>", width = "100%")`
:::
:::

```{r mat_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  mat_icb_objs$map,
  mat_icb_objs$chart,
  ncol = 2
)
```
```{r mat_icb_vis_download}
get_download_button(title = "Download chart data", data = mat_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. This is the latest available population statistics at the time of publication. To align with women who may benefit from maternity exemption certificates, population figures are based on the female population aged 15-45. However, these population statistics only cover a single year and will include women who are not pregnant and have not recently given birth.

Bedfordshire, Luton and Milton Keynes ICB had the highest estimated number of certificates issued per 10,000 population. In 2023/24, an 636 certificates were issued per 10,000 population in this ICB.

Cheshire and Merseyside ICB had the lowest estimated number of certificates issued per 10,000 population in 2023/24. This ICB had an 408 certificates per 10,000 population.

------------------------------------------------------------------------

### 2.3 Medical exemption certificate

`r infoBox_border("",text = "Medical exemption certifications exempt the holders from NHS prescription charges only. Patients with any of the qualifying conditions can apply for a certificate. But it is the certificate that provides exemption - not the condition.

Patients can apply for a medical exemption certificate if they have either:

* a permanent fistula (for example, caecostomy, colostomy, laryngostomy or ileostomy) which needs continuous surgical dressing or an appliance
* a form of hypoadrenalism (for example, Addison’s Disease) for which specific substitution therapy is essential
* diabetes insipidus and other forms of hypopituitarism
* diabetes mellitus, except where treatment is by diet alone
* hypoparathyroidism
* myasthenia gravis
* myxoedema (hypothyroidism which needs thyroid hormone replacement)
* epilepsy which needs continuous anticonvulsive therapy
* a continuing physical disability which means you cannot go out without the help of another person
* cancer and are undergoing treatment for either:
  * cancer
  * the effects of cancer
  * the effects of cancer treatment

Medical exemption certificates are usually valid for 5 years. A GP or doctor can provide the application form to qualifying patients."
, width = "100%")`

#### Figure X: Number of medical exemption certificate applications received 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 495,000 applications received in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Applications dropped during 2020/21 and 2021/22 but have since increased</b>", width = "100%")`
:::
:::

```{r medex_applications_chart}
med_application_objs$chart
get_download_button(title = "Download chart data", data = med_application_objs$chart_data, filename = "hwhc_figure_x")
```

Applications for medical exemption certificates dipped in 2020/21 and 2021/22 but has since increased to the highest level in recent years. 

In 2023/24 there were 495,000 applications received, up from 478,000 in 2022/23.

Applications are authorised by a healthcare professional, who confirms eligibility. Therefore, only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

#### Figure X: Number of medical exemption certificates issued 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 475,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Certificates issued dropped during 2020/21 and 2021/22 but have since increased</b>", width = "100%")`
:::
:::

```{r medex_issued_chart}
med_issued_objs$chart
get_download_button(title = "Download chart data", data = med_issued_objs$chart_data, filename = "hwhc_figure_x")
```

The number of issued medical exemption certificates dipped in 2020/2021 but has since increased to the highest level in recent years.

In 2023/24 there were 475,000 medical exemption certificates issued, up from 453,000 in 2022/23.

Medical exemption certificates are usually valid for a period of 5 years. The number of active certificates in any year will be notably higher than the number of certificates issued that year.

#### Figure X: Estimated number of active medical exemption certificates 2019/20 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 2,530,000 active medical exemption certificates in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Numbers are relatively stable with around 2,500,000 active certificates each year</b>", width = "100%")`
:::
:::

```{r medex_active_charts}
med_active_objs$chart
get_download_button(title = "Download chart data", data = med_active_objs$chart_data, filename = "hwhc_figure_x")
```

The number of active certificates is only available from 2019/20 due to retention schedules for certificates prior to this period. It should be noted that active certificates are based solely on the start and expiration date of the issued certificates. This does not consider if the certificate holder was using the certificate or not.

Over the past five years there have been around 2,500,000 medical exemption certificates active each year. The 2,530,000 active certificates in 2023/24 is a 2% decrease from the 2,580,000. Although the number of certificates issued has increased over the past couple of years, the drop in certificates issued from 2020/21 will be reflected in the number of active certificates for the next couple of years.

#### Figure X: Number of medical exemption certificates issued, split by age of recipient (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of certificates issued rises as the age of recipient rises</b>", width = "100%")`
:::
::: col-md-6
`r infoBox_no_border("", text = "<b>134,000 certificates were issued to people aged 55-59 in 2023/24</b>", width = "100%")`
:::
:::

```{r medex_age_chart}
med_age_objs$chart
get_download_button(title = "Download chart data", data = med_age_objs$chart_data, filename = "hwhc_figure_x")
```

There is a strong relationship between age and medical exemption certificates. The number of certificates issued steadily rises in line with age, which may be expected due to long-term conditions and multi-morbidity being more prevalent in older people. In 2023/24 the number of issued certificates ranged from 8,950 in the 15-19 age group, to 134,000 in the 55-59 age group.

As a medical exemption certificate only entitles the holder to free NHS prescriptions, it is not required after the age of 60, when NHS prescriptions are automatically free.

#### Figure X: Number of medical exemption certificates issued, split by deprivation profile (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The least deprived areas of the country had the lowest number of medical exemption certificates issued in 2023/34</b>", width = "100%")`
:::
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of medical exemption certificates issued in the most deprived quintile was lower than the next two quintiles with less deprivation</b>", width = "100%")`
:::
:::

```{r medex_age_profile_chart}
med_imd_objs$chart
get_download_button(title = "Download chart data", data = med_imd_objs$chart_data, filename = "hwhc_figure_x")
```

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

There is a slight correlation between deprivation of an area and the number of medical exemption certificates issued. Excluding the most deprived area, as deprivation decreases so does the number of issued certificates.

The number of certificates issued in the most deprived areas was slightly lower than that for the next two, less deprived, groups.


#### Figure X: Number of medical exemption certificates issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24 figures ranged from 224 certificates per 10,000 population in Greater Manchester ICB to 318 per 10,000 population in Lincolnshire ICB</b>", width = "100%")`
:::
:::

```{r med_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  med_icb_objs$map,
  med_icb_objs$chart,
  ncol = 2
)
```
```{r med_icb_vis_download}
get_download_button(title = "Download chart data", data = med_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24, and will include patients who do not qualify for support via the medical exemption certificate. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to a ICB or age group from the available data.

There is not a large variation between ICBs. Values range from 224 certificates per 10,000 population in Greater Manchester ICB to 318 per 10,000 population in Lincolnshire ICB.

------------------------------------------------------------------------

### 2.4 NHS Prescription Prepayment Certificate (PPC)

`r infoBox_border("",text = "A PPC can help provide savings for people who pay for NHS Prescriptions. The certificate covers all a patient's NHS prescription charges for a set price. Patients will save money if they need 4 or more items in 3 months or 12 or more items in 12 months.

PPCs can be purchased in-person at some pharmacies or direct via the [PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc). Applicants can choose between a 3-month and 12-month certificate.

Patients receiving prescriptions for hormone replacement therapy (HRT) may benefit more from the Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC)."
, width = "100%")`

#### Figure X: Number of PPC applications received 2018/19 to 2023/24, split by PPC type {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 3,150,000 applications received in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Applications for PPC have increased each and every year</b>", width = "100%")`
:::
:::

```{r ppc_applications_chart}
ppc_application_objs$chart
get_download_button(title = "Download chart data", data = ppc_application_objs$chart_data, filename = "hwhc_figure_x")
```

Applications for 12-month PPCs have shown a consistent year-on-year increase. In 2023/24 there were 2,160,000 applications received, up from 2,060,000 in 2022/23 which is a 5% increase. The number of applications for 12-month certificates is more than double the number of applications for 3-month certificates. 

The year-on-year increase for 3-month certificate applications is not as high, and applications dropped in 2020/21. In 2023/24 there were 992,000 applications for 3-month certificates, up 2% from the 973,000 applications received in 2022/23.

The number of applications for 12-month PPCs is roughly double the number of applications for 3-month PPCs, with this rate gradually increasing each year. If a PPC is required for the full year, a 12-month certificate will cost less than 3-month certificates for the same period.

Not all applications will result in a certificate being issued. Certificates will not be issued if there are issues with the initial payment for the certificate and in some cases the applicant may drop out for other reasons.

#### Figure X: Number of PPCs issued 2018/19 to 2023/24, split by PPC type {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 3,000,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The number of PPC certificates issued has increased each and every year</b>", width = "100%")`
:::
:::

```{r ppc_issued_chart}
ppc_issued_objs$chart
get_download_button(title = "Download chart data", data = ppc_issued_objs$chart_data, filename = "hwhc_figure_x")
```

The number of 12-month PPC certificates issued has shown a consistent year-on-year increase. In 2023/24 there were 2,060,000 certificates issued, up from 1,980,000 in 2022/23 which is a 4% increase. The number of 12-month certificates issued is more than double the number of 3-month certificates issued. 

The year-on-year increase for 3-month certificates issued is not as high, and the number of certificates issued dropped in 2020/21. In 2023/24 there were 936,000 certificates issued, up 1% from the 923,000 certificates issued in 2022/23.

#### Figure X: Number of PPCs issued by age of recipient, split by PPC type (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of certificates issued rises as the age of the recipient rises</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24, there were 782,000 certificates issued to people aged 55-59</b>", width = "100%")`
:::
:::

```{r ppc_age_profile_chart}
ppc_age_objs$chart
get_download_button(title = "Download chart data", data = ppc_age_objs$chart_data, filename = "hwhc_figure_x")
```

There is a strong relationship between age PPC certificates issued. 

In 2023/24, the number of 12-month certificates issued ranged from 19,900 for people 15-19 to 527,000 for people aged 55-59. There is a similar pattern for 3-month certificates but numbers are lower, ranging from 13,500 in the 15-19 age group to 255,000 for people aged 55-59.

As a PPC only provides support for NHS prescription charges, it is not required after the age of 60, when NHS prescriptions are automatically free. For applicants who choose to auto-renew their certificate, the certificate will not be renewed for people turning 60 within 9 months of their existing certificate expiring.

The 12-month certificate is more popular across all age bands. However, the ratio of 3-month to 12-month certificates generally increases in line with age. In the 15-19 age group the number of 12-month PPCs is only 1.5 times the number of 3-month PPCs. In the 50-54 age group this rate rises to 2.8.

#### Figure X: Number of PPCs issued by deprivation profile, split by PPC type (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The number of certificates issued are lower in the most and least deprived areas of the country</b>", width = "100%")`
:::
:::

```{r ppc_imd_chart}
ppc_imd_objs$chart
get_download_button(title = "Download chart data", data = ppc_imd_objs$chart_data, filename = "hwhc_figure_x")
```

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

There is not a strong relationship between deprivation and number of PPCs issued. However, numbers are lower in the most and least deprived areas of the country.

#### Figure X: Number of PPCs issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24 the five London ICBs were the bottom five in terms of certificates issued per 10000 population</b>", width = "100%")`
:::
:::

```{r ppc_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  ppc_icb_objs$map, 
  ppc_icb_objs$chart, 
  ncol = 2
)
```
```{r ppc_icb_vis_download}
get_download_button(title = "Download chart data", data = ppc_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to a ICB or age group from the available data.

The map and chart are based on the total number of 3-month and 12-month certificates issued in 2023/24. Figures for individual certificate types are included in the accompanying data.

In 2023/24 the five London ICBs were the bottom five in terms of certificates issued per 10000 population. The lowest figure was for North Central London ICB, with 853 certificates issued per 10,000 population in 2023/24. 

Lincolnshire ICB had the highest figure in 2023/24, with 2,721 certificates issued per 10,000 population.

------------------------------------------------------------------------

### 2.5 NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC)

`r infoBox_border("",text = "The HRT PPC was introduced in April 2023 to provide support with the prescription costs for medication used to treat symptoms of the menopause. 

A HRT PPC covers all a patient's NHS prescription charges, for certain HRT medicines, for a set price. HRT PPCs are valid for 12 months and patients will save money if they need 3 or more HRT prescription items in 12 months.

The HRT PCC can be used for any of the applicable HRT medicines, regardless of why they are prescribed. The HRT PPC webpages include a [list of the medicines covered by a HRT PPC](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc/medicines-covered-hrt-ppc).

HRT PPCs can be purchased in-person at some pharmacies or direct via the [HRT PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc).

Patients receiving prescriptions for other medicines, include HRT medicines not covered by the HRT PPC, may benefit more from a 3 or 12 month Prescription Prepayment Certificate (PPC) that covers all NHS prescriptions."
, width = "100%")`

#### Figure X: Number of HRT PPC applications received (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In the first year it was available, 617,000 applications were received for HRT PPCS</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>There were 143,000 applications received in the first month</b>", width = "100%")`
:::
:::

```{r hrtppc_applications_chart}
hrt_application_objs$chart
get_download_button(title = "Download chart data", data = hrt_application_objs$chart_data, filename = "hwhc_figure_x")
```

Following the launch of the HRT PPC in April 2023, there were 143,000 applications received in the first month alone. The number of applications then declined each month but have started to show signs 
of growth near the end of the year. 

As this is the first year the HRT PPC has been available it is difficult to assume any meaningful trends from the available data. A HRT PCC is valid for 12 months, and therefore any certificates issued in 2023/24 will not need replacing until 2024/25.

#### Figure X: Number of HRT PPCs issued (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In the first year it was available, 566,000 HRT PPCs were issued.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Almost half of the certificates issued in March 2024 were set to start in April 2024</b>", width = "100%")`
:::
:::

```{r hrtppc_issued_chart}
highcharter::hw_grid(
  hrt_issued_objs$chart |> 
    highcharter::hc_title(text = 'Number of HRT PPCs issued'), 
  ch_hrt_issued_postdate |> 
    highcharter::hc_title(text = 'Proportion of issued certificates dated to start the following month'), 
  ncol = 2
)
```
```{r hrtppc_issued_chart_download}
get_download_button(title = "Download chart data", data = hrt_issued_objs$chart_data, filename = "hwhc_figure_x")
```

The data for issued certificates is very similar to the application data. This is to be expected as there is no qualifying criteria for HRT PPC. However certificates may not be issued if there are any issues with with the application that are not resolved.

There were 566,000 HRT PPCs issued in 2023/24. Although numbers decreased after a very strong launch of the service, numbers have started to increase near the end of the year when existing certificates will be nearing expiration.

The 131,000 certificates issued in the first month represent 23% of all HRT PPC certificates issued in 2023/24.

The number of certificates issued leveled out near the end of the near with a notable jump in March 2024 when 37,600 certificates were issued. However, almost half of the certificates issued in March 2024 had a certificate start date in April 2024 which may suggest a large number of these were customers renewing expiring certificates. 

#### Figure X: Number of HRT PPCs issued by age of recipient (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Over 92% of HRT PPCs were issued to recipients aged 45 and over</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The age group with the most HRT PPCs issued was 50-54</b>", width = "100%")`
:::
:::

```{r hrtppc_age_chart, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  hrt_age_objs$chart |> 
    highcharter::hc_title(text = 'Number of HRT PPCs issued (2023/24)'), 
  hrt_age_base_pop_objs$chart |> 
    highcharter::hc_title(text = 'Estimated number of patients receiving HRT PPC qualifying medication (2023/24)'), 
  ncol = 2
)
```
```{r hrtppc_age_chart_download}
get_download_button(title = "Download chart data", data = hrt_age_objs$chart_data, filename = "hwhc_figure_x")
```

Over 92% of HRT PPCs were issued to people aged 45 to 59. This may be expected as the menopause usually affects women between the ages of 45 and 55. Certificates peaked at the 50 to 54 age group, with 226,000 HRT PPCs issued to this age group.

The distribution of HRT PPCs by age is similar to the distribution of the estimated number of patients receiving the HRT qualifying medication on NHS prescriptions in 2023/24.

#### Figure X: Number of HRT PPCs issued by deprivation profile (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of HRT PPCs issued rises as deprivation decreases</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Only 37,600 certificates were issued in the most deprived areas in 2023/24</b>", width = "100%")`
:::
:::

```{r hrtppc_imd_chart, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  hrt_imd_objs$chart |> 
    highcharter::hc_title(text = 'Number of HRT PPCs issued (2023/24)'), 
  hrt_imd_base_pop_objs$chart |> 
    highcharter::hc_title(text = 'Estimated number of patients receiving HRT PPC qualifying medication (2023/24)'), 
  ncol = 2
)
```
```{r hrtppc_imd_chart_download}
get_download_button(title = "Download chart data", data = hrt_imd_objs$chart_data, filename = "hwhc_figure_x")
```

There is a strong relationship between deprivation and the number of HRT PPCs issued in 2023/24. There were only 37,600 certificates issued in the most deprived areas. There were 189,000 certificates issued in the least deprived areas.

The distribution of HRT PPCs by deprivation is similar to the distribution of the estimated number of patients receiving the HRT qualifying medication on NHS prescriptions in 2023/24. However, the relationship between deprivation and issued certificates is a bit more pronounced for HRT PPCs issued, with less certificates issued to the most deprived areas. In the most deprived areas, people may be more likely to qualifying for other methods of support with health costs.

#### Figure X: Number of HRT PPCs issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>ICBs in the South West show higher rates of HRT PPCs per 10,000 than other areas of the country.</b>", width = "100%")`
:::
:::

```{r hrt_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  hrt_icb_objs$map, 
  hrt_icb_objs$chart, 
  ncol = 2
)
```
```{r hrt_icb_vis_download}
get_download_button(title = "Download chart data", data = hrt_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions for drugs covered by the HRT PPC during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to a ICB or age group from the available data.

Values range from 2774 certificates per 10,000 population in Black Country ICB to 4175 per 10,000 population in Cornwall and Isles of Scilly ICB.

In 2023/24, 5 of the 7 ICBs in the South West region appeared in the top 10 for number of HRT PPCs issued per 10,000 population. ICBs in the Midlands, North West and North East and Yorkshire regions had lower rates of certificates issued per 10,000 population.

------------------------------------------------------------------------

### 2.6 NHS tax credit exemption certificates

`r infoBox_border("",text = "NHS tax credit exemption certificates provide support for:

* free NHS prescriptions
* free NHS dental treatment
* free NHS sight tests
* help with the cost of glasses or contact lenses
* free NHS wigs and fabric supports
* help with the cost of travel to receive NHS treatment

There is no application process as HM Revenue and Customs will identify who is eligible. Certificates are issued to people if the annual income used to calculate Tax Credits is £15,276 or less and they receive either:

* Child Tax Credit
* Working Tax Credit and Child Tax Credit paid together
* Working Tax Credit including a disability or severe disability element

Certificates are valid for 7 months, and issued every 6 months in April and September."
, width = "100%")`

#### Figure X: Number of NHS tax credit exemption certificates issued 2018/19 to 2023/24 {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 1,150,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Certificates issued have been consistently dropping each year</b>", width = "100%")`
:::
:::

```{r tax_issued_chart}
tax_issued_objs$chart
get_download_button(title = "Download chart data", data = tax_issued_objs$chart_data, filename = "hwhc_figure_x")
```

The number of issued Tax Credit Exemption Certificates has been dropping year on year from 3,740,000 in 2018/19 to 1,150,000 in 2023/24. This strong decline may be linked to the Universal Credit rollout as people are migrated to new benefit schemes. The number of certificates issued each year has declined by between 22% and 26% since 2019/2020.

#### Figure X: Number of NHS tax credit exemption certificates issued, split by age band (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of certificates issued peaks at age 40-44</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>77% of certificates were issued to people aged 35 to 54</b>", width = "100%")`
:::
:::

```{r tax_age_chart}
tax_age_objs$chart
get_download_button(title = "Download chart data", data = tax_age_objs$chart_data, filename = "hwhc_figure_x")
```

The number of Tax Credit Exemption Certificates issued in 2023/24 peaked at ages 40-44 with 268,000 certificates issued.

Over three quarters of all certificates were issued to people aged between 35 and 54.

Although numbers are relatively small, some certificates are issued to people aged 60 and over as Tax Credit Exemption Certificate also provides support with some health costs which do not have an age exemption.

#### Figure X: Change in age profile of NHS tax credit exemption certificates issued (2018/19 to 2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2018/19, over 55% of NHS tax credit exemption certificates were issued to people under 40 years old</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>By 2023/24 this had dropped to just over 30% of issued certificates.</b>", width = "100%")`
:::
:::

```{r tax_age_profile_chart}
ch_tax_age_trend
get_download_button(title = "Download chart data", data = dl_tax_age_trend, filename = "hwhc_figure_x")
```

The age profile of Tax Credit Exemption Certificates have changed considerably over the past few years. In 2018/19 over 55% of certificates holders were under 40 years old, but as of 2023/24 this group represents only 30% of certificates issued. Following the introduction of Universal Credit, new applicants will receive support via Universal Support rather than tax credits.

#### Figure X: Number of NHS tax credit exemption certificates issued by deprivation profile (2023/24) {.toc-ignore}


::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The number of certificates issued to the most deprived areas is almost six times the number issued in the least deprived areas</b>", width = "100%")`
:::
:::

```{r tax_imd_chart}
tax_imd_objs$chart
get_download_button(title = "Download chart data", data = tax_imd_objs$chart_data, filename = "hwhc_figure_x")
```

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure. These indices of deprivation are specific to England and therefore can only be applied to certificates that can be assigned to an English postcode.

There is a strong relationship between deprivation and number of NHS tax credit exemption certificates issued. This is to be expected as this is a means-tested benefit. In 2023/24, there were 394,000 certificates issued in the most deprived areas. In the least deprived areas only 67,700 certificates were issued in 2023/24.

#### Figure X: Number of NHS tax credit exemption certificates issued per 10,000 population, by ICB (2023/24) {.toc-ignore}

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24 figures ranged from 85 certificates per 10,000 population in Bath and North East Somerset, Swindon and Wiltshire ICB to 510 per 10,000 population in North East London ICB</b>", width = "100%")`
:::
:::

```{r tax_icb_vis, figures-side, fig.show="hold", out.width="50%"}
highcharter::hw_grid(
  tax_icb_objs$map, 
  tax_icb_objs$chart, 
  ncol = 2
)
```
```{r tax_icb_vis_download}
get_download_button(title = "Download chart data", data = tax_icb_objs$chart_data, filename = "hwhc_figure_x")
```

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. This is the latest available population statistics at the time of publication. To align with NHS tax credit exemption certificates, population figures are based on the population aged 16 and over.

There are a small number of ICBs where the number of certificates per 10,000 population is much higher than the rest of the country.

In 2023/24, North East London ICB had 510 certificates issued per 10,000 population. This was closely followed by 507 certificates per 10,000 population in Birmingham and Solihull ICB. Numbers then drop to 386 certificates per 10,000 population in North Central London ICB.

In 2023/24, Bath and North East Somerset, Swindon and Wiltshire ICB had the lowest rate, with only 85 certificates issued per 10,000 population.

------------------------------------------------------------------------

## 3 About these statistics

### 3.1 Statistical Disclosure Control (SDC) {.toc-ignore}

Statistical disclosure control has been applied to these statistics. Volume figures have been redacted in the supporting summary tables if they relate to fewer than 5 items. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

### 3.2. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases rounding has been to two significant figures in the key findings and information boxes and three significant figures in the main text. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 3.3. Index of deprivation {#about_imd .toc-ignore}

The [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) have been used to provide a measure of deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals' living conditions.

The reported IMD quintile, where 1 is the most deprived and 5 the least deprived, is derived from the postcode currently held for an applicant. There are a small number of records each year that we have reported as having an unknown IMD quintile. These are applicants with no fixed abode or with british forces addresses, or where we have been unable to match the applicant postcode to an English postcode in the [XXX citation needed](https://geoportal.statistics.gov.uk/datasets/national-statistics-postcode-lookup-may-2022/about).

IMD quintiles are calculated by ranking census lower-layer super output areas (LSOAs) from most deprived to least deprived and dividing them into equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

------------------------------------------------------------------------

## 4 Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:
**Email**:
[accessibility\@nhsbsa.nhs.uk](mailto:accessibility@nhsbsa.nhs.uk){.email}

**Phone**: 0191 203 5318

[Find out about call
charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We'll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the 'Contact us' section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

------------------------------------------------------------------------

## 5 Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](%60r%20config$stats_survey_link%60) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:
**Email:**
[statistics\@nhsbsa.nhs.uk](mailto:statistics@nhsbsa.nhs.uk){.email}

**You can also write to us at:**
NHSBSA - Statistics\
NHS Business Services Authority\
Stella House\
Goldcrest Way\
Newburn Riverside\
Newcastle upon Tyne\
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>
