---
title: "`r config$publication_top_name`"
pagetitle: NHS Help with Health Costs
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '2'
  html_document:
    anchor_sections: no
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<html lang="en">

```{r setup, include=FALSE}

# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)


```


<div class = "quartoheader", id = "fixed-header">
  <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
</div>

<!-- JS to handle hiding the header on scroll -->
<script>
  document.addEventListener("scroll", function() {
    var header = document.getElementById("fixed-header");
    var toc = document.getElementById("TOC");
    var BackTopButton = document.getElementById("backtop-button");
    var scrollPosition = window.scrollY || document.documentElement.scrollTop;
  if(scrollPosition > 88) {
    header.style.top = "-200px";
    toc.style.setProperty('margin-top', '24px', 'important');
    BackTopButton.style.setProperty('visibility', 'visible', 'important');
  } else
  {
    header.style.top = "0";
    toc.style.setProperty('margin-top', '140px', 'important');
    BackTopButton.style.setProperty('visibility', 'hidden', 'important');
  }
}
);
</script>

<!-- JS to handle back to top button -->
<script>
       function BackTop() {
        document.body.scrollTop = 0;
        document.documentElement.scrollTop = 0;
    }
</script>

<button onclick="BackTop()" id="backtop-button" style="opacity: 0.7; pointer-events: initial; visibility: visible;"><span aria-hidden="true">▲</span>Back to top</button>


<main>

# `r config$publication_sub_name` {.toc-ignore}

`r paste("Published", config$publication_date)`

## Key findings 2024/25

`r infoBox_border("Background",text = "The NHSBSA administers a range of services to help people with their health costs, including help with:

* NHS prescription costs
* NHS dental costs
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Full details of these schemes are available on the [NHS Help with Health Costs webpage](https://www.nhsbsa.nhs.uk/nhs-help-health-costs).

Six different certificate schemes are available, with different levels of help with health costs:

* [Maternity exemption certificates](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/maternity-exemption-certificates)
* [Medical exemption certificates](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/medical-exemption-certificates)
* [NHS Tax Credit Exemption Certificates](https://www.nhsbsa.nhs.uk/exemption-certificates/nhs-tax-credit-exemption-certificates)
* [NHS Low Income Scheme](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme)
* [NHS Prescription Prepayment certificate     (PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc)
* [NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT     PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc).

"
, width = "100%")`

### Maternity exemption certificates {.toc-ignore}

The number of certificates issued has risen each year since 2020/21. There were 574,000 maternity exemption certificates issued in 2024/25.

Generally, less deprived areas of the country received fewer maternity exemption certificates than more deprived areas. The number of certificates issued was highest in 30-34 year olds, accounting for 36% of all certificates issued.

### Medical exemption certificates {.toc-ignore}

The number of certificates issued has risen over the past 3 years, following a drop between 2019/20 and 2021/22. There were 484,000 medical exemption certificates issued in 2024/25.

The number of certificates issued is higher in older age groups than in younger age groups.

### NHS Tax Credit Exemption Certificates {.toc-ignore}

As part of the Universal Credit rollout, people have been transferred away from tax credits and onto new benefit schemes. Tax Credit payments made by HMRC ended on 5 April 2025, so the NHS Tax Credit exemption certificate scheme is no longer issuing certificates. The number of certificates issued declined for a number of years before the end of the service. In 2024/25, 359,000 certificates were issued, compared to 3.7 million in 2018/19.

The age profile of recipients has changed in the last few years of the scheme. In 2018/19, 55% of certificates were issued to people aged under 40 years old but in 2024/25 this dropped to 27% of certificates issued.

### NHS Low Income Scheme {.toc-ignore}

The total number of certificates issued in 2024/25 fell to 242,000. This was due to the number of HC2 certificates issued decreasing for the second year in a row, to 154,000 certificates. There were 87,900 HC3 certificates, a slight increase from 2023/24.

Certificates were most commonly issued to people from the youngest age groups of 15-19 and 20-24, and the oldest age group of 65+. More certificates were issued to people from more deprived areas.

### NHS Prescription Prepayment Certificate (PPC) {.toc-ignore}

The total number of PPCs purchased has increased every year since 2020/21. There were 2.2 million 12-month certificates issued and 984,000 3-month certificates issued in 2024/25.

The number of PPCs issued increases in older age groups. There is no clear pattern by deprivation in England, with the number of certificates issued in the most and least deprived areas slightly lower than in other areas.

### NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC) {.toc-ignore}

The HRT PPC was introduced on 1 April 2023. There were 542,000 HRT PPCs issued in 2024/25, a decrease from 2023/24. However, from July 2024 to March 2025 the number of certificates issued was higher than in the same month in the previous year.

The number of certificates issued was higher in less deprived areas. 

Integrated Care Boards (ICBs) in the South and South West regions generally have higher rates of HRT PPCs issued.


------------------------------------------------------------------------

## 1. Things you should know

### 1.1. Scope {.toc-ignore}

These statistics provide estimates of volumes related to the 6 certificate schemes. Data covers the financial years 2018/2019 to 2024/25, where the financial year runs from 1 April to 31 March. The time period covered by these statistics includes the Coronavirus (COVID-19) pandemic period. It is worth noting that during this time period there was a reduction in face-to-face primary care consultations, referrals, new prescriptions and diagnoses due to the COVID-19 pandemic and national lockdowns.

These statistics are based on information from applications and data captured during certificate processing activities. Therefore, where information has not been supplied or captured accurately, this may impact the quality of the statistics. The NHSBSA takes measures to minimise potential errors including monthly quality checks based on a random sample of certificates issued. As of May 2024, the 12-month rolling accuracy results for NHS Help with Health Costs service areas were above 98%. 

Data in this publication may differ from other sources due to differences in how certificates are assigned to an issue year or issue month. It is possible for small differences in data to happen if the underlying data is extracted on different dates. 

Figures are presented based on the number of applications and issued certificates. Figures should not be interpreted as numbers of people as a unique individual could be represented multiple times.

The NHS Low Income Scheme and NHS Tax Credit Exemption Certificates provide support to people across the United Kingdom whilst the other NHS Help with Health Costs services administered by the NHSBSA are typically focused on support for residents in England. Within this report, figures represent all activity, with additional breakdowns by country available in the supporting datasets.

Integrated Care Board (ICB) areas have been used within this narrative to allow comparisons between areas. Applicants are aligned to an ICB area based on their postcode and mappings available using the [National Statistics Postcode Lookup (NSPL) - August 2024](https://geoportal.statistics.gov.uk/datasets/c5afedb9204a47e99559a4880feddcb1/about).

### 1.2. Purpose {.toc-ignore}

This publication can have a wide range of uses, including:

* informing government or local NHS policy
* monitoring the uptake of NHS Help with Health Costs schemes, including by ICB and by Index of Multiple Deprivation (IMD) quintiles
* allowing scrutiny by the public and stakeholders

The data used to produce this summary narrative is available as data downloads for each figure. The data is also available in the supporting summary tables on the [HWHC 2024/25 release webpage](https://www.nhsbsa.nhs.uk/statistical-collections/help-health-costs/help-health-costs-england-202425). A data dashboard has been produced to allow users to explore the data in more detail. 

------------------------------------------------------------------------

## 2. Results and commentary

### 2.1. Maternity exemption certificate

`r infoBox_border("",text = "Maternity exemption certificates exempt the holders from NHS prescription charges. They can also be used as proof of entitlement to free NHS dental treatment whilst pregnant.

Maternity exemption certificates are available to anybody who is currently pregnant or has given birth in the last 12 months. The certificate is valid until 12 months from the baby's due date or the birth of the baby.

Applications are submitted via a midwife, doctor or health visitor who confirms eligibility. Both paper and digital applications are available. 

Only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

As a maternity exemption certificate provides support with NHS prescription charges, support is generally applicable to applicants from England only." 
, width = "100%")`

#### Number of maternity exemption certificates issued 2018/19 to 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 574,000 certificates issued in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Issued certificates increased by 4% from 2023/24 to 2024/25.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 1: The number of certificates issued has increased each year since 2020/21

Number of maternity exemption certificates issued from 2018/19 to 2024/25

```{r matex_issued_chart}
mat_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = mat_issued_objs$chart_data, filename = "hwhc_figure_1")
```

##### Table

###### Table 1: The number of certificates issued has increased each year since 2020/21

Number of maternity exemption certificates issued from 2018/19 to 2024/25

```{r matex_issued_table}
mat_issued_objs$table
get_download_button(title = "Download table data", data = mat_issued_objs$chart_data, filename = "hwhc_table_1")
```

</div>

The number of maternity exemption certificates issued decreased in 2020/2021 but has increased in each of the following years. In 2024/25 there were 574,000 maternity exemption certificates issued, 4% more than the 552,000 certificates issued in 2023/24. This was a smaller percentage change than the 12% increase between 2022/23 and 2023/24. 

#### Number of maternity exemption certificates issued, split by age band of applicant 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Numbers are highest in the 30-34 age band, with 206,000 certificates issued to this age group.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Distribution of issued certificates by age follows a similar pattern to live birth statistics.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 2: Certificates issued were highest in the 30-34 age group

Number of maternity exemption certificates issued in 2024/25 by age of applicant

```{r matex_age_chart}
mat_age_objs$chart
get_download_button(title = "Download chart data", data = mat_age_objs$chart_data, filename = "hwhc_figure_2")
```

##### Supporting Information (Live Births)
###### Figure 2 (supporting information): Live births are highest in the 30-34 age group

Number of live births in 2023, by age of mother

```{r matex_age_support_chart}
ch_births_age
get_download_button(title = "Download chart data", data = ch_births_age_data, filename = "hwhc_figure_2")
```

Source: [2023 Live birth figures from the Office for National Statistics](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/birthsinenglandandwalesbirthregistrations)

##### Table
###### Table 2: Certificates issued were highest in the 30-34 age group

Number of maternity exemption certificates issued in 2024/25 by age of applicant

Note: In Table 2, ONS live births data for under 20s are included alongside the 15-19 age band for certificates issued. However, the under 20 age band in the ONS lives births data will include live births for those aged under 15. 

```{r matex_age_table}
mat_age_objs$table
get_download_button(title = "Download table data", data = mat_age_objs$chart_data, filename = "hwhc_table_2")
```

</div>

The distribution of maternity exemption certificates by age is highest at 30-34, with 206,000 certificates issued to this age group in 2024/25. 

There is no publicly available data that accurately shows the number of women eligible for maternity exemptions in a specific time period. However, the age distribution is very similar to the latest available live birth statistics published by the Office for National Statistics (ONS) for 2023. The live births data does not cover the same time period as the 2024/25 financial year of the certificates issued, and is included for context only.


#### Number of maternity exemption certificates issued, split by Index of Multiple Deprivation (IMD) quintile 2024/25 {.tabset}

<div class = "tabset">

##### Chart

###### Figure 3: The number of certificates issued was lowest in the least deprived areas
 
Number of maternity exemption certificates issued in 2024/25 by IMD quintile of recipient

```{r mat_imd_chart}
mat_imd_objs$chart
get_download_button(title = "Download chart data", data = mat_imd_objs$chart_data, filename = "hwhc_figure_3")
```

##### Supporting Information (Live Births)
###### Figure 3 (supporting information): The number of live births decreases as deprivation decreases

Number of live births in 2023, by IMD quintile of mother

```{r mat_imd_support_chart}
ch_births_imd
get_download_button(title = "Download chart data", data = mat_imd_objs$chart_data, filename = "hwhc_figure_3")
```

Source: [2023 Live birth figures from the Office for National Statistics](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/birthsinenglandandwalesbirthregistrations)

##### Table

###### Table 3: The number of certificates issued is lowest in the least deprived areas

Number of maternity exemption certificates issued in 2024/25 by IMD quintile of recipient

```{r mat_imd_table}
mat_imd_objs$table
get_download_button(title = "Download table data", data = mat_imd_objs$chart_data, filename = "hwhc_table_3")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. These indices of deprivation are specific to England so they have only been applied to certificates that can be assigned to an English address. You can find more information on the English Indices of Deprivation in section 3 of this summary.

The number of certificates issued is generally lower in less deprived areas, with the least deprived areas having the fewest number of certificates issued. Quintile 1 and 2 have similar numbers of certificates issued, whereas in the latest available live birth statistics published by the ONS, the number of births is higher in quintile 1. In the most deprived areas, women may be more likely to qualify for other income-related benefits that offer the same support as a maternity exemption certificate.

#### Number of maternity exemption certificates issued per 10,000 population, by ICB 2024/25 {.tabset}

<div class = "tabset">

##### Map

###### Figure 4 (map): Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2024/25 ICB areas, per 10,000 female population aged 15 to 45

```{r mat_icb_map}
mat_icb_objs$map
get_download_button(title = "Download map data", data = mat_icb_objs$chart_data, filename = "hwhc_figure_4")
```

##### Chart

###### Figure 4 (chart): Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2024/25, by ICB area, per 10,000 female population aged 15 to 45

```{r mat_icb_chart}
mat_icb_objs$chart
get_download_button(title = "Download chart data", data = mat_icb_objs$chart_data, filename = "hwhc_figure_4")
```

##### Table

###### Table 4: Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2024/25, by ICB area, per 10,000 female population aged 15 to 45

```{r mat_icb_table}
mat_icb_objs$table
get_download_button(title = "Download table data", data = mat_icb_objs$chart_data, filename = "hwhc_table_4")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates use the [2022 mid-year population estimates by ICB](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates) published by the ONS, the latest available at time of publication. Population estimates used have been limited to the female population aged 15 to 45, to align with women who may be most likely to benefit from maternity exemption certificates. There is currently no publicly available data on the number of women who would qualify for support via the maternity exemption certificate at a point in time.

Bedfordshire, Luton and Milton Keynes ICB had the highest estimated number of certificates issued per 10,000 population in 2024/25. In this ICB an estimated 657 certificates were issued per 10,000 population.

The ICB with the lowest estimated number of certificates issued per 10,000 population in 2024/25 was Cheshire and Merseyside ICB. This ICB had an estimated 429 certificates per 10,000 population.

------------------------------------------------------------------------

### 2.2. Medical exemption certificate

`r infoBox_border("",text = "Medical exemption certificates exempt the holders from NHS prescription charges only. Patients with any of the qualifying conditions can apply for a certificate. It is the certificate that provides exemption - not the condition.

Patients can apply for a medical exemption certificate if they have either:

* a permanent fistula (for example, caecostomy, colostomy, laryngostomy or ileostomy) which needs continuous surgical dressing or an appliance
* a form of hypoadrenalism (for example, Addison’s Disease) for which specific substitution therapy is essential
* diabetes insipidus and other forms of hypopituitarism
* diabetes mellitus, except where treatment is by diet alone
* hypoparathyroidism
* myasthenia gravis
* myxoedema (hypothyroidism which needs thyroid hormone replacement)
* epilepsy which needs continuous anticonvulsive therapy
* a continuing physical disability which means you cannot go out without the help of another person
* cancer and are undergoing treatment for either:
  * cancer
  * the effects of cancer
  * the effects of cancer treatment

Applications are submitted via a GP or doctor who confirms eligibility. Medical exemption certificates are usually valid for 5 years. Certificates will not be automatically renewed upon expiration, with certificate holders needing to reapply.

Only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

As a medical exemption certificate provides support with NHS prescription charges, support is generally applicable to applicants from England aged between 15 and 59."
, width = "100%")`

#### Number of medical exemption certificates issued 2018/19 to 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 484,000 certificates issued in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Certificates issued dropped during 2020/21 and 2021/22 but have since increased.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 5: The number of certificates issued has increased each year since 2020/21

Number of medical exemption certificates issued from 2018/19 to 2024/25

```{r medex_issued_chart}
med_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = med_issued_objs$chart_data, filename = "hwhc_figure_5")
```

##### Table

###### Table 5: The number of certificates issued has increased each year since 2020/21

Number of medical exemption certificates issued from 2018/19 to 2024/25

```{r medex_issued_table}
med_issued_objs$table
get_download_button(title = "Download table data", data = med_issued_objs$chart_data, filename = "hwhc_table_5")
```

</div>

The number of issued medical exemption certificates dipped in 2020/2021 but has since increased to the highest level in recent years.

In 2024/25 there were 484,000 medical exemption certificates issued, a 1.7% increase from 475,000 in 2023/24.

#### Number of medical exemption certificates issued, split by age of applicant 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of certificates issued increases with age.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>133,000 certificates were issued to people aged 55-59 in 2024/25.</b>", width = "100%")`
:::
:::


##### Chart

###### Figure 6: The number of certificates issued is higher in older age bands

Number of medical exemption certificates issued in 2024/25 by age of applicant

```{r medex_age_chart}
med_age_objs$chart
get_download_button(title = "Download chart data", data = med_age_objs$chart_data, filename = "hwhc_figure_6")
```

##### Table

###### Table 6: The number of certificates issued is higher in older age bands

Number of medical exemption certificates issued in 2024/25 by age of applicant

```{r medex_age_table}
med_age_objs$table
get_download_button(title = "Download table data", data = med_age_objs$chart_data, filename = "hwhc_table_6")
```

</div>

There is a strong relationship between age and medical exemption certificates. The number of certificates issued increases as age increases, which may be expected due to long-term conditions and multi-morbidity being more prevalent in older people. In 2024/25, the number of issued certificates in the 15-19 age group was 9,280, rising to 133,000 in the 55-59 age group.

A medical exemption certificate only entitles the holder to free NHS prescriptions. Therefore, it is not required after the age of 60 when NHS prescriptions are automatically free.

#### Number of medical exemption certificates issued, split by IMD quintile 2024/25 {.tabset}

<div class = "tabset">

##### Chart

###### Figure 7: Certificates are issued less frequently to people in the least deprived areas

Number of medical exemption certificates issued in 2024/25 by IMD quintile

```{r medex_imd_chart}
med_imd_objs$chart
get_download_button(title = "Download chart data", data = med_imd_objs$chart_data, filename = "hwhc_figure_7")
```

##### Table

###### Table 7: Certificates are issued less frequently to people in the least deprived areas

Number of medical exemption certificates issued in 2024/25 by IMD quintile

```{r medex_imd_table}
med_imd_objs$table
get_download_button(title = "Download table data", data = med_imd_objs$chart_data, filename = "hwhc_table_7")
```

</div>

The least deprived areas had the fewest certificates issued in 2024/25. Quintile 2 had the highest number of certificates issued. It should be noted that the number of issued certificates has not been adjusted by IMD population estimates, as current population estimates by IMD were unavailable.

The number of certificates issued in quintile 2, the second most deprived areas, was slightly higher than in quintile 1, the most deprived areas. In the most deprived areas people may be more likely to qualify for other income-related benefits that offer the same support as a medical exemption certificate.

#### Number of medical exemption certificates issued per 10,000 population, by ICB 2024/25 {.tabset}

<div class = "tabset">

##### Map

###### Figure 8 (map): Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2024/25 by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_map}
med_icb_objs$map
get_download_button(title = "Download map data", data = med_icb_objs$chart_data, filename = "hwhc_figure_8")
```

##### Chart

###### Figure 8 (chart): Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2024/25 by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_chart}
med_icb_objs$chart
get_download_button(title = "Download chart data", data = med_icb_objs$chart_data, filename = "hwhc_figure_8")
```

##### Table

###### Table 8: Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2024/25 by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_table}
med_icb_objs$table
get_download_button(title = "Download table data", data = med_icb_objs$chart_data, filename = "hwhc_table_8")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients aged 16-59 who were identified as receiving NHS prescriptions during 2024/25. The population estimates include all patients regardless of whether or not they paid prescription charges in 2024/25. The estimates also include patients who do not qualify for support via the medical exemption certificate. These population estimates may exclude some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to an ICB or age group from the available data.

Overall, the number of certificates issued does not vary much between ICBs. Cheshire and Merseyside ICB had the lowest number of medical exemption certificates issued per 10,000 population, at 216, while Lincolnshire ICB had the highest, at 312.

------------------------------------------------------------------------

### 2.3. NHS Tax Credit Exemption Certificates

`r infoBox_border("",text = "Tax Credit payments made by HM Revenue and Customs (HMRC) ended on 5 April 2025, so this scheme is no longer issuing certificates. NHS Tax Credit Exemption Certificates provided support for:

* free NHS prescriptions
* free NHS dental treatment
* free NHS sight tests
* help with the cost of glasses or contact lenses
* free NHS wigs and fabric supports
* help with the cost of travel to receive NHS treatment

There was no application process as HMRC identified who was eligible. Certificates were issued to people if their annual income used to calculate Tax Credits was £15,276 or less and they received either:

* Child Tax Credit
* Working Tax Credit and Child Tax Credit paid together
* Working Tax Credit including a disability or severe disability element

Certificates are valid for up to 7 months, and certificates issued before the end of the scheme will remain valid until they run out.

The majority of activity is for certificates issued to people in England, however certificates are also processed for people from Scotland, Wales and Northern Ireland.

People receiving tax credits were moved to Universal Credit. After the end of the tax credit scheme, they must meet the Universal Credit conditions of the Low Income Scheme to be entitled to full help with health costs."
, width = "100%")`

#### Number of NHS Tax Credit Exemption Certificates issued 2018/19 to 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 359,000 certificates issued in 2024/25, the last full year of the Tax Credit exemption scheme.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The number of certificates issued fell every year from 2018/19 to 2024/25.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 9: The number of certificates issued has dropped year on year

Number of NHS Tax Credit Exemption Certificates issued from 2018/19 to 2024/25

```{r tax_issued_chart}
tax_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = tax_issued_objs$chart_data, filename = "hwhc_figure_9")
```

##### Table

###### Table 9: The number of certificates issued has dropped year on year

Number of NHS Tax Credit Exemption Certificates issued by financial year from 2018/19 to 2024/25

```{r tax_issued_table}
tax_issued_objs$table
get_download_button(title = "Download table data", data = tax_issued_objs$chart_data, filename = "hwhc_table_9")
```

</div>

The number of NHS Tax Credit Exemption Certificates issued has decreased every year since 2018/19, from 3.7 million in 2018/19 to 359,000 in 2024/25. In 2024/25, 69% fewer certificates were issued than the 1.2 million certificates in 2023/24. 

As part of the Universal Credit roll-out people are being transferred away from tax credits and onto new benefit schemes. This has reduced the number of people in receipt of tax credits and the scheme stopped issuing new certificates after the last tax credit payments were made on 5 April 2025.

#### Number of NHS Tax Credit Exemption Certificates issued, split by age band of applicant 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>77% of certificates were issued to people aged 35 to 54.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The age group with the highest number of certificates issued was people aged 40 to 44.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 10: The number of certificates issued peaks at age 40-44

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, split by age of applicant

```{r tax_age_chart}
tax_age_objs$chart
get_download_button(title = "Download chart data", data = tax_age_objs$chart_data, filename = "hwhc_figure_10")
```

##### Table

###### Table 10: The number of certificates issued peaks at age 40-44

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, split by age of applicant

```{r tax_age_table}
tax_age_objs$table
get_download_button(title = "Download table data", data = tax_age_objs$chart_data, filename = "hwhc_table_10")
```

</div>

The number of NHS Tax Credit Exemption Certificates issued in 2024/25 was highest in the 40-44 age group, with 83,400 certificates issued to this group.

Most certificates were issued to people aged between 35 and 54, at 77% of the total certificates issued for 2024/25.

Although prescription items have an age exemption, some other health costs do not. Therefore, certificates were also available to people aged 60 and over. There were 26,400 certificates issued to this group in 2024/25.

#### Number of NHS Tax Credit Exemption Certificates issued by IMD quintile 2024/25 {.tabset}

<div class = "tabset">

##### Chart

###### Figure 11: Certificates are most commonly issued to the most deprived areas

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, split by IMD quintile

```{r tax_imd_chart}
tax_imd_objs$chart
get_download_button(title = "Download chart data", data = tax_imd_objs$chart_data, filename = "hwhc_figure_11")
```

##### Table

###### Table 11: Certificates are most commonly issued to the most deprived areas

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, split by IMD quintile

```{r tax_imd_table}
tax_imd_objs$table
get_download_button(title = "Download table data", data = tax_imd_objs$chart_data, filename = "hwhc_table_11")
```

</div>

Tax credit certificates were a means-tested benefit and more certificates were issued to those in the most deprived areas compared to the least deprived areas. There were 124,000 certificates issued in the most deprived areas in 2024/25, almost 7 times more than the 18,600 certificates in the least deprived areas.

The indices of deprivation used in this publication are specific to England so can only be applied to certificates assigned to an English address.

#### Number of NHS Tax Credit Exemption Certificates issued per 10,000 population, by ICB 2024/25 {.tabset}

<div class = "tabset">

##### Map

###### Figure 12 (map): Certificates issued per 10,000 population varies by ICB area

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r tax_icb_map}
tax_icb_objs$map
get_download_button(title = "Download map data", data = tax_icb_objs$chart_data, filename = "hwhc_figure_12")
```

##### Chart

###### Figure 12 (chart): Certificates issued per 10,000 population varies by ICB area

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r tax_icb_chart}
tax_icb_objs$chart
get_download_button(title = "Download chart data", data = tax_icb_objs$chart_data, filename = "hwhc_figure_12")
```

##### Table

###### Table 12: Certificates issued per 10,000 population varies by ICB area

Number of NHS Tax Credit Exemption Certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r tax_icb_table}
tax_icb_objs$table
get_download_button(title = "Download table data", data = tax_icb_objs$chart_data, filename = "hwhc_table_12")
```

</div>

ICB areas have been used to allow comparisons between areas, though ICBs are not responsible for NHS Tax Credit Exemption Certificates. Applicants have been associated with an ICB area based on their postcode, which is only possible for people within England.

Population estimates use mid-year population estimates published by the [ONS for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). The 2022 population estimates used for this report are the latest available at the time of publication. To align with NHS Tax Credit Exemption Certificates, population figures are based on the population aged 16 and over.

The number of certificates issued per 10,000 population varies by ICB. In 2024/25, the highest number of certificates issued per 10,000 population was 216 for Birmingham and Solihull ICB, while Bath and North East Somerset, Swindon and Wiltshire ICB had the lowest, at 19 per 10,000 population.

------------------------------------------------------------------------

### 2.4. NHS Low Income Scheme

`r infoBox_border("",text = "The NHS Low Income Scheme provides income related help. Support is available for people who do not receive help with health costs via an existing exemption.

NHS Low Income Scheme exemption provides help with:

* NHS prescriptions
* NHS dental treatment
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Applicants either complete a paper HC1 form or apply via the [online application for NHS Low Income Scheme](https://apply-for-help-with-nhs-costs.nhsbsa.nhs.uk/apply-online). Two levels of support are available from the NHS Low Income Scheme. Depending on the outcome of the assessment, applicants could receive:

* an [HC2 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc2-certificates-full-help-health-costs) for full help with health costs, including free NHS prescriptions
* an [HC3 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc3-certificates-limited-help-health-costs) for limited help with health costs, which will show how much the holder has to pay towards health costs

The amount of support provided is based on weekly income, necessary outgoings, plus any savings or investments. Issued certificates may also cover the applicant's partner and dependent children under 20 years old.

The majority of certificates are issued for a year, but some are issued for periods of between 1 day and 5 years, depending on circumstances.

The majority of activity is for applicants from England, however applications are also processed for applicants from Scotland and Wales.

These statistics exclude any figures for certificates issued to asylum seekers assisted by UK Visas and Immigration (UKVI). These individuals are issued certificates automatically as part of the support they receive and are not handled through the regular application and assessment process."
, width = "100%")`

#### Number of NHS Low Income Scheme certificates issued by certificate type from 2018/19 to 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>The number of HC2 certificates issued decreased in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The number of HC3 certificates increased in 2024/25.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 13: Almost two-thirds of certificates issued in 2024/25 provided full help with health costs

Number of NHS Low Income Scheme certificates issued, by certificate type, from 2018/19 to 2024/25

```{r lis_issued_chart}
lis_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = lis_issued_objs$chart_data, filename = "hwhc_figure_13")
```

##### Table

###### Table 13: Almost two-thirds of certificates issued in 2024/25 provided full help with health costs

Number of NHS Low Income Scheme certificates issued, by certificate type, from 2018/19 to 2024/25

```{r lis_issued_table}
lis_issued_objs$table
get_download_button(title = "Download table data", data = lis_issued_objs$chart_data, filename = "hwhc_table_13")
```

</div>

In 2024/25, there were 242,000 certificates issued for the Low Income Scheme. Of this total, 154,000 were HC2 certificates and 87,900 were HC3 certificates. The total number of certificates decreased compared to 2023/24, but only HC2 certificates decreased. HC3 certificates increased slightly by 1%.

These figures do not include any applications that are ongoing or were abandoned before completion, as certificates are only issued to applicants once the assessment has been completed. The figures also exclude completed applications where no certificate was issued to the applicant.

#### Number of NHS Low Income Scheme certificates issued, split by certificate type and applicant age band in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>More HC2 certificates were issued to applicants aged 65 and over than any other age group in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, applicants aged 20 to 24 were issued the most HC3 certificates.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 14: More than half of certificates issued were to people aged under 25 or 65 and over

Number of NHS Low Income Scheme certificates issued in 2024/25, by certificate type and applicant age band

```{r lis_age_chart}
lis_age_objs$chart
get_download_button(title = "Download chart data", data = lis_age_objs$chart_data, filename = "hwhc_figure_14")
```

##### Table

###### Table 14: More than half of certificates issued were to people aged under 25 or 65 and over

Number of NHS Low Income Scheme certificates issued in 2024/25, by certificate type and applicant age band

```{r lis_age_table}
lis_age_objs$table
get_download_button(title = "Download table data", data = lis_age_objs$chart_data, filename = "hwhc_table_14")
```

</div>

NHS Low Income Scheme certificates cover the applicant, partners and any dependent children, but age information is only captured for the lead applicant.

The oldest age group and 2 youngest age groups received more certificates than other ages. In 2024/25, 48,400 certificates went to people aged 65 and over, 20% of the total. There were 48,800 certificates issued to applicants aged 20-24 and 31,900 certificates issued to applicants aged 19 and under.

#### Number of NHS Low Income Scheme certificates issued, split by certificate type and IMD quintile in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25 there were 68,800 certificates issued to people in the most deprived areas.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, the second most deprived quintile received almost as many HC2 certificates as the most deprived quintile.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 15: The number of certificates issued in the most deprived areas was more than 3 times higher than the least deprived areas

Number of NHS Low Income Scheme certificates issued in 2024/25, split by certificate type and IMD quintile

```{r lis_imd_chart}
lis_imd_objs$chart
get_download_button(title = "Download chart data", data = lis_imd_objs$chart_data, filename = "hwhc_figure_15")
```

##### Table

###### Table 15: The number of certificates issued in the most deprived areas was more than 3 times higher than the least deprived areas

Number of NHS Low Income Scheme certificates issued in 2024/25, split by certificate type and IMD quintile

```{r lis_imd_table}
lis_imd_objs$table
get_download_button(title = "Download table data", data = lis_imd_objs$chart_data, filename = "hwhc_table_15")
```

</div>

There were 68,800 certificates issued in the most deprived areas, more than 3 times higher than the 21,600 certificates issued in the least deprived areas.

The 68,800 certificates issued in quintile 1, the most deprived areas, was similar to the 62,200 certificates issued in quintile 2, the second most deprived areas.

In the most deprived areas, people may be more likely to qualify for other income-related benefits that offer some of the same support as available through the NHS Low Income Scheme.

IMD quintiles only show a generalised level of deprivation for geographic areas, so will not reflect the circumstances of all people in those areas. Certificates are means-tested and are still issued to applicants from less deprived areas if they meet the conditions. In 2024/25, 13,700 HC2 certificates and 7,940 HC3 certificates went to applicants from quintile 5, the least deprived areas.

The indices of deprivation used in this publication are specific to England, so can only be applied to certificates assigned to an English address.

#### Number of NHS Low Income Scheme certificates issued per 10,000 population, by ICB in 2024/25 {.tabset}

<div class = "tabset">

The map, chart and table are based on the total combined number of HC2 and HC3 certificates issued in 2024/25. Figures for individual certificate types are included in the supporting data.

##### Map

###### Figure 16 (map): Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r lis_icb_map}
lis_icb_objs$map
get_download_button(title = "Download map data", data = lis_icb_objs$chart_data, filename = "hwhc_figure_16")
```

##### Chart

###### Figure 16 (chart): Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r lis_icb_chart}
lis_icb_objs$chart
get_download_button(title = "Download chart data", data = lis_icb_objs$chart_data, filename = "hwhc_figure_16")
```

##### Table

###### Table 16: Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2024/25, by ICB area, per 10,000 population aged 16 and over

```{r lis_icb_table}
lis_icb_objs$table
get_download_button(title = "Download table data", data = lis_icb_objs$chart_data, filename = "hwhc_table_16")
```

</div>

ICB areas have been used to allow comparisons between areas, however ICBs are not responsible for applications to the NHS Low Income Scheme. Applicants have been aligned to an ICB area based on their postcode, which is only possible for applicants within England.

Population estimates use figures published by the [ONS for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report, the latest available at the time of publication. To align with the people who may benefit from the NHS Low Income Scheme, population figures have been limited to people aged 16 and over.

In 2024/25, an estimated 82 certificates were issued per 10,000 population in North West London ICB, the highest of all ICBs. Out of the 4 ICBs with the most certificates issued per 10,000 population, 3 were North London ICBs.

The lowest estimated number of certificates issued per 10,000 population in 2024/25 was 28 in Bath and North East Somerset, Swindon and Wiltshire ICB.

------------------------------------------------------------------------

### 2.5. NHS Prescription Prepayment Certificate (PPC)

`r infoBox_border("",text = "A PPC can help provide savings for people who pay for their NHS Prescriptions. The certificate covers all NHS prescription charges for a set price. Patients could save money if they need 4 or more items in 3 months or 12 or more items in 12 months.

PPCs can be purchased in-person at some pharmacies or direct via the [PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc). Applicants can choose between a 3-month and 12-month certificate.

Only a small proportion of applications would not end with a certificate being issued. Certificates will not be issued if there are issues with the initial payment for the certificate and in some cases the applicant may drop out for other reasons. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

Patients receiving prescriptions for hormone replacement therapy (HRT) may benefit more from the Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC).

As a PPC provides support with NHS prescription charges, support is generally only applicable to applicants from England aged between 16 and 59."
, width = "100%")`

#### Number of PPCs issued from 2018/19 to 2024/25, split by PPC type {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 3.2 million PPCs issued in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The number of 12-month PPCs issued and 3-month PPCs issued increased in 2024/25.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 17: The total number of PPCs issued has increased each year since 2020/21

Number of PPCs issued from 2018/19 to 2024/25, split by PPC type

```{r ppc_issued_chart}
ppc_issued_objs$chart
get_download_button(title = "Download chart data", data = ppc_issued_objs$chart_data, filename = "hwhc_figure_17")
```

##### Table

###### Table 17: The total number of PPCs issued has increased each year since 2020/21

Number of PPCs issued from 2018/19 to 2024/25, split by PPC type

```{r ppc_issued_table}
ppc_issued_objs$table
get_download_button(title = "Download table data", data = ppc_issued_objs$chart_data, filename = "hwhc_table_17")
```

</div>

The number of 12-month PPCs issued has increased every year since 2020/21. There were 2.2 million 12-month certificates issued in 2024/25, a 6% increase from 2.1 million in 2023/24. The number of 12-month certificates issued is more than twice as many as the number of 3-month certificates issued. 

The number of 3-month certificates decreased in 2020/21 but has increased every year since. In 2024/25, there were 984,000 3-month certificates issued, 5% more than in 2023/24. This was a similar percentage increase as for 12-month certificates between 2023/24 and 2024/25.

#### Number of PPCs issued by age of applicant, split by PPC type in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, the number of certificates issued increased with age for 12-month and 3-month PPCs.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, there were a total of 820,000 certificates issued to people aged 55-59.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 18: The number of certificates issued rises as the age of the applicant rises

Number of PPCs issued in 2024/25, split by PPC type and age of applicant

```{r ppc_age_chart}
ppc_age_objs$chart
get_download_button(title = "Download chart data", data = ppc_age_objs$chart_data, filename = "hwhc_figure_18")
```

##### Table

###### Table 18: The number of certificates issued rises as the age of the applicant rises

Number of PPCs issued in 2024/25, split by PPC type and age of applicant

```{r ppc_age_table}
ppc_age_objs$table
get_download_button(title = "Download table data", data = ppc_age_objs$chart_data, filename = "hwhc_table_18")
```

</div>

In 2024/25, the number of 12-month certificates issued for people aged 15-19 was 20,900, compared to 557,000 for people aged 55-59. While the number of 3-month certificates issued is lower, it shows a similar pattern across age groups. There were 14,100 certificates issued in 2024/25 to people aged 15-19, and 264,000 certificates issued to people aged 55-59.

A PPC only provides support for NHS prescription charges. Therefore, it is not required after the age of 60, when NHS prescriptions are automatically free. For applicants who choose to auto-renew their certificate, the certificate will not be renewed if they turn 60 within 9 months of their existing certificate expiring.

All age groups have more people receiving a 12-month certificate, and the ratio of 12-month to 3-month certificates increases across older age groups. In the 15-19 age group, 1.5 times more 12-month PPCs were issued compared to the number of 3-month PPCs. This rises to almost 3 times more 12-month PPCs than 3-month PPCs in the 50-54 age group, then drops to 2.1 times more 12-month than 3-month PPCs for the 55-59 age group.

#### Number of PPCs issued by deprivation profile, split by PPC type in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
:::

##### Chart

###### Figure 19: Slightly fewer certificates are issued in the most and least deprived areas

Number of PPCs issued in 2024/25, split by PPC type and IMD quintile

```{r ppc_imd_chart}
ppc_imd_objs$chart
get_download_button(title = "Download chart data", data = ppc_imd_objs$chart_data, filename = "hwhc_figure_19")
```

##### Table

###### Table 19: Slightly fewer certificates are issued in the most and least deprived areas

Number of PPCs issued in 2024/25, split by PPC type and IMD quintile

```{r ppc_imd_table}
ppc_imd_objs$table
get_download_button(title = "Download table data", data = ppc_imd_objs$chart_data, filename = "hwhc_table_19")
```

</div>

There is not an obvious pattern of the number of PPCs issued across IMD quintiles. The most and least deprived areas of the country have less certificates issued than the other 3 IMD quintiles. People may be more likely to qualify for other income-related benefits that provide access to free NHS prescriptions in the most deprived areas.

#### Number of PPCs issued per 10,000 population, by ICB in 2024/25 {.tabset}

<div class = "tabset">

The map, chart and table are based on the total number of 3-month and 12-month certificates issued in 2024/25. Figures for individual certificate types are included in the accompanying data.

##### Map

###### Figure 20 (map): Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2024/25, by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_map}
ppc_icb_objs$map
get_download_button(title = "Download map data", data = ppc_icb_objs$chart_data, filename = "hwhc_figure_20")
```

##### Chart

###### Figure 20 (chart): Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2024/25, by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_chart}
ppc_icb_objs$chart
get_download_button(title = "Download chart data", data = ppc_icb_objs$chart_data, filename = "hwhc_figure_20")
```

##### Table

###### Table 20: Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2024/25, by ICB area, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_table}
ppc_icb_objs$table
get_download_button(title = "Download table data", data = ppc_icb_objs$chart_data, filename = "hwhc_table_20")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions during 2024/25. The population estimates include all patients regardless of whether or not they paid prescription charges in 2024/25. These population estimates may exclude patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to an ICB or age group from the available data.

In 2024/25, the ICBs with the 5 lowest numbers of certificates issued per 10,000 population were all London ICBs. The ICB with the fewest number of certificates issued was North Central London ICB, with 888 certificates issued per 10,000 population. 

Lincolnshire ICB had 2,740 certificates issued per 10,000 population, the highest of all ICBs in 2024/25.

------------------------------------------------------------------------

### 2.6. HRT PPC

`r infoBox_border("",text = "The HRT PPC was introduced on 1 April 2023 to provide support with the prescription costs for medication used to treat symptoms of the menopause. 

The HRT PPC covers all NHS prescription charges for certain HRT medicines for a set price. HRT PPCs are valid for 12 months and patients could save money if they need 3 or more HRT prescription items in 12 months.

The HRT PPC can be used for any of the applicable HRT medicines, regardless of why they are prescribed. The HRT PPC webpages include a [list of the medicines covered by a HRT PPC](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc/medicines-covered-hrt-ppc).

HRT PPCs can be purchased in-person at some pharmacies or direct via the [HRT PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc).

Only a small proportion of applications would not end with a certificate being issued. Certificates will not be issued if there are issues with the initial payment for the certificate and in some cases the applicant may drop out for other reasons. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

If a patient's medication is not covered, or they also receive prescriptions for other medication, they may benefit more from a 3 or 12 month PPC that covers all NHS prescriptions.

As an HRT PPC provides support with NHS prescription charges, support is generally only applicable to applicants from England aged between 16 and 59."
, width = "100%")`

#### Number of HRT PPCs issued in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>542,000 HRT PPCs were issued in 2024/25.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Fewer HRT PPCs were issued in 2024/25 than in 2023/24.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 21: July 2024 had the highest number of certificates issued in 2024/25

Number of HRT PPCs issued by month during 2024/25

```{r hrt_issued_chart}
hrt_issued_objs$chart
get_download_button(title = "Download chart data", data = hrt_issued_objs$chart_data, filename = "hwhc_figure_21")
```

##### Supporting Information

###### Figure 21 (post-dated certificates): March 2024 and 2025 had a sharp increase in the proportion of certificates post-dated to start the following month

Proportion of issued HRT PPCs post-dated to start the following month, by month of certificate issue

```{r hrt_issued_support}
ch_hrt_issued_postdate
get_download_button(title = "Download chart data", data = hrt_issued_objs$chart_data, filename = "hwhc_figure_21")
```

##### Table

###### Table 21: July 2024 had the highest number of certificates issued in 2024/25

Number of HRT PPCs issued by month during 2024/25

```{r hrt_issued_table}
hrt_issued_objs$table
get_download_button(title = "Download table data", data = hrt_issued_objs$chart_data, filename = "hwhc_table_21")
```

</div>

The total number of certificates fell after a strong first year of the service. There were 542,000 HRT PPCs issued in 2024/25, a 4% decrease from 566,000 certificates issued in 2023/24. However, from July 2024 to March 2025, the number of certificates issued each month was higher than the same months in the previous year. 

In 2024/25, the 4 months at the start of the financial year had higher numbers of certificates issued by month. As certificates are valid for 12 months, certificates issued early in 2023/24 began to expire in early 2024/25. As with 2023/24, the number of certificates issued rose in March, the last month of the 2024/25 financial year. Almost a third of certificates issued in March 2025 had a certificate start date in April 2025. This may be due to customers renewing expiring certificates.

July 2024 had 60,400 certificates issued, the highest by month in 2024/25 with 11% of the total for the financial year. The distribution of certificates issued by month varied less than in 2023/24, when 23% of certificates were issued in April 2023 when the service began.

#### Number of HRT PPCs issued by age of applicant in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>91% of HRT PPCs were issued to people aged 45 and over.</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The age group with the most HRT PPCs issued was 50-54.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 22: Most people issued HRT PPCs were aged 45 and over

Number of HRT PPCs issued in 2024/25, split by age of applicant


```{r hrt_age_chart}
hrt_age_objs$chart
get_download_button(title = "Download chart data", data = hrt_age_objs$chart_data, filename = "hwhc_figure_22")
```

##### Supporting Information

###### Figure 22 (NHS prescribing): Most patients receiving HRT PPC eligible medication were aged 45 and over

Estimated number of patients receiving NHS prescribing in 2024/25 of medication qualifying for support by HRT PPC, split by patient age band

```{r hrt_age_support}
hrt_age_base_pop_objs$chart
get_download_button(title = "Download chart data", data = hrt_age_objs$chart_data, filename = "hwhc_figure_22")
```

##### Table

###### Table 22: Most people issued HRT PPCs were aged 45 and over

Number of HRT PPCs issued in 2024/25, split by age of applicant

```{r hrt_age_table}
hrt_age_objs$table
get_download_button(title = "Download table data", data = hrt_age_objs$chart_data, filename = "hwhc_table_22")
```

</div>

In 2024/25, 91% of HRT PPCs were issued to people aged 45 to 59. As menopause usually affects women between the ages of 45 and 55, higher numbers of HRT PPCs in these age groups may be expected. The highest number of certificates issued was in the 50-54 age group, at 212,000.

There is a similar distribution of HRT PPCs by age when compared to the distribution of the estimated number of patients receiving HRT qualifying medication on NHS prescriptions in 2024/25.

#### Number of HRT PPCs issued by deprivation profile in 2024/25 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, only 36,400 HRT PPCs were issued in the most deprived areas.</b>", width = "100%")`
:::


::: col-md-6
`r infoBox_no_border("", text = "<b>In 2024/25, 181,000 HRT PPCs were issued in the least deprived areas.</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 23: The number of HRT PPCs issued rises as deprivation decreases

Number of HRT PPCs issued in 2024/25, by IMD quintile

```{r hrt_imd_chart}
hrt_imd_objs$chart
get_download_button(title = "Download chart data", data = hrt_imd_objs$chart_data, filename = "hwhc_figure_23")
```

##### Supporting Information

###### Figure 23 (eligible patients): Patients in the least deprived areas are more likely to receive prescribing of HRT medication

Estimated number of patients receiving NHS prescribing in 2024/25 of medication qualifying for support by HRT PPC, split by IMD quintile

```{r hrt_imd_support}
hrt_imd_base_pop_objs_test$chart
get_download_button(title = "Download chart data", data = hrt_imd_objs$chart_data, filename = "hwhc_figure_23")
```

##### Table

###### Table 23: The number of HRT PPCs issued rises as deprivation decreases

Number of HRT PPCs issued in 2024/25, by IMD quintile

```{r hrt_imd_table}
hrt_imd_objs$table
get_download_button(title = "Download table data", data = hrt_imd_objs$chart_data, filename = "hwhc_table_23")
```

</div>

There is a strong relationship between deprivation and the number of HRT PPCs issued in 2024/25. There were 181,000 certificates issued in the least deprived areas, almost 5 times more than the 36,400 certificates issued in the most deprived areas.

The distribution of HRT PPCs by deprivation is similar to the distribution of the estimated number of patients receiving HRT qualifying medication on NHS prescriptions in 2024/25. However, the difference between quintile 1 and quintile 2 is more pronounced for HRT PPCs. In 2024/25 there were 19% fewer estimated HRT patients in quintile 1 than quintile 2, but 49% fewer HRT PPCs. In the most deprived areas, people may be more likely to qualify for other methods of help with health costs.

It should be noted that the number of issued certificates has not been adjusted by IMD population estimates, as current population estimates by IMD were unavailable.

#### Number of HRT PPCs issued per 10,000 population, by ICB in 2024/25 {.tabset}

<div class = "tabset">

##### Map

###### Figure 24 (map): Some ICBs in the South and South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2024/25 by ICB area per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_map}
hrt_icb_objs$map
get_download_button(title = "Download map data", data = hrt_icb_objs$chart_data, filename = "hwhc_figure_24")
```

##### Chart

###### Figure 24 (chart): Some ICBs in the South and South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2024/25 by ICB area per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_chart}
hrt_icb_objs$chart
get_download_button(title = "Download chart data", data = hrt_icb_objs$chart_data, filename = "hwhc_figure_24")
```

##### Table

###### Table 24: Some ICBs in the South and South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2024/25 by ICB area per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_table}
hrt_icb_objs$table
get_download_button(title = "Download table data", data = hrt_icb_objs$chart_data, filename = "hwhc_table_24")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients, aged 16-59, identified as receiving NHS prescriptions for drugs covered by the HRT PPC during 2024/25. The population estimates include all patients regardless of whether or not they paid prescription charges in 2024/25. These population estimates may exclude patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to an ICB or age group from the available data.

The number of certificates issued per 10,000 population varied by ICB in 2024/25, from 2,430 in Black Country ICB to 3,750 in South West London ICB.

Overall, ICBs in the Soth and South West regions of England tend to have a higher number of HRT PPCs issued per 10,000 than ICBs in other regions.

------------------------------------------------------------------------

## 3. About these statistics

### 3.1. Statistical Disclosure Control (SDC) {.toc-ignore}

Statistical disclosure control has been applied to these statistics. Volume figures have been redacted in the supporting summary tables if they relate to fewer than 5 items or patients. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

### 3.2. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases, rounding has been to 2 significant figures in the key findings and information boxes and 3 significant figures in the main text. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 3.3. Index of deprivation {#about_imd .toc-ignore}

The [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) have been used to provide a measure of deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals' living conditions.

The reported IMD quintile, where 1 is the most deprived and 5 the least deprived, is derived from the postcode currently held for an applicant. There are a small number of records each year that we have reported as IMD quintile being "not available". These are applicants with no fixed abode or with British forces addresses, or where we have been unable to match the applicant postcode to an English postcode in the [National Statistics Postcode Lookup (NSPL) - August 2024](https://geoportal.statistics.gov.uk/datasets/c5afedb9204a47e99559a4880feddcb1/about).

IMD quintiles are calculated by ranking census lower-layer super output areas (LSOAs) from most deprived to least deprived and dividing them into equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

### 3.4. ICB classification {#about_icb .toc-ignore}

ICB areas have been used to allow comparisons between areas. Applicants have been associated with an ICB area based on their postcode using the [National Statistics Postcode Lookup (NSPL) - August 2024](https://geoportal.statistics.gov.uk/datasets/c5afedb9204a47e99559a4880feddcb1/about). ICBs are only available within England, so this geography classification is only available for applicants with a valid English postcode.

### 3.5. Age classification {#about_age .toc-ignore}

Reporting by age band uses the applicants age at the time of application. Figures have been grouped into 5 year age bands. The age is calculated using the captured date of birth information for the applicant. If the date of birth is not available the age will be reported as "not available".

Patients under 16 do not usually require NHSBSA issued exemption certificates as NHS treatment is free for this age group.

Patients aged 60 and over are automatically entitled to free NHS prescriptions and therefore would not require maternity exemption certificates, medical exemption certificates, or prescription prepayment certificates.

------------------------------------------------------------------------

## 4. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: [accessibility\@nhsbsa.nhs.uk](mailto:accessibility@nhsbsa.nhs.uk){.email}

**Phone**: 0191 203 5318

[Find out about call
charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We'll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the 'Contact us' section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

------------------------------------------------------------------------

## 5. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:**
[statistics\@nhsbsa.nhs.uk](mailto:statistics@nhsbsa.nhs.uk){.email}

**You can also write to us at:**

NHSBSA - Statistics\
NHS Business Services Authority\
Stella House\
Goldcrest Way\
Newburn Riverside\
Newcastle upon Tyne\
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>

