---
title: "`r config$publication_top_name`"
pagetitle: NHS Help with Health Costs
always_allow_html: true
output:
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    anchor_sections: false
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{=html}
<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>
```
```{=html}
<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
```

<html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<div class = "quartoheader", id = "fixed-header">
        <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
    </div>

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date`

## Key findings 2023/24

`r infoBox_border("Background",text = "The NHSBSA administer a range of services to help people with their health costs, including help with:

* NHS prescription costs
* NHS dental costs
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

Full details of these schemes are available on the [NHS Help with Health Costs webpage](https://www.nhsbsa.nhs.uk/nhs-help-health-costs).

Six different certificate schemes are available, with different levels of help with health costs:

* [Maternity exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/maternity-exemption-certificates)
* [Medical exemption certificate](https://www.nhsbsa.nhs.uk/check-if-you-have-nhs-exemption/medical-exemption-certificates)
* [NHS Tax credit exemption certificate](https://www.nhsbsa.nhs.uk/exemption-certificates/nhs-tax-credit-exemption-certificates)
* [NHS Low Income Scheme (LIS)](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme)
* [NHS Prescription Prepayment certificate     (PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc)
* [NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT     PPC)](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)"
, width = "100%")`

### Maternity exemption certificates {.toc-ignore}

The number of certificates issued has risen each year since 2020/21. There were 552,000 maternity exemption certificates issued in 2023/24.

The most deprived areas of the country have the highest number of births. However, the most deprived areas do not receive the highest number of maternity exemption certificates.

### Medical exemption certificates {.toc-ignore}

The number of certificates issued has risen following a drop between 2019/20 and 2021/22. There were 475,000 medical exemption certificates issued in 2023/24.

The number of certificates issued typically rises with age.

### NHS tax credit exemption certificates {.toc-ignore}

The number of NHS tax credit exemption certificates has been declining for a number of years. As part of the Universal Credit rollout people are being migrated away from tax credits and onto new benefit schemes. In 2023/24, 1.2 million certificates were issued, compared to 3.7 million in 2018/19.

The age profile of recipients has changed over the past few years. In 2018/19, 55% of certificates were issued to people aged under 40 years old but in 2023/24 this has dropped to 30% of issued certificates.

### NHS Low Income Scheme {.toc-ignore}

The number of applications and issued certificates in 2023/24 was a slight decrease from 2022/23 but higher than the 2 years prior. There were 161,000 HC2 certificates and 87,000 HC3 certificates issued in 2023/24.

Certificates were most commonly issued to people from the younger (15-24) and older (65+) age groups, and to people from more deprived areas.

### NHS Prescription Prepayment Certificate (PPC) {.toc-ignore}

The number of PPCs purchased has increased in each of the past 5 years. In 2023/24 there were 2.1 million 12-month certificates issued and 936,000 3-month certificates issued.

The number of certificates issued typically rises with age.

The number of PPCs issued is slightly lower in the most and least deprived areas of England. 

### NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC) {.toc-ignore}

The HRT PPC was launched in April 2023, with 131,000 certificates being issued in the first month. This represents 23% of the 566,000 certificates issued in 2023/24.

The number of certificates issued was higher in less deprived areas. There was also a higher number of patients in the least deprived areas receiving HRT medicines.

The South West region has some of the largest rates of HRT PPCs issued.


------------------------------------------------------------------------

## 1. Things you should know

### 1.1. Scope {.toc-ignore}

These statistics provide estimates of volumes related to these certificate schemes. Data covers the financial years 2018/2019 to 2023/24, where the financial year runs from April 1 to March 31.

These statistics are based on information from applications and data captured during certificate processing activities. Therefore, where information has not been supplied or captured accurately, this may impact the quality of the statistics. The NHSBSA take measures to minimise potential errors including monthly quality checks based on a random sample of certificates issued. As of May 2024 the 12-month rolling accuracy results for NHS Help with Health Costs service areas were above 98%.  

Figures are presented based on the number of applications and issued certificates. Figures should not be interpreted as numbers of people as a unique individual could be represented multiple times.

The NHS Low Income Scheme and NHS tax credit exemption certificate provide certificates to people across the United Kingdom whilst the other NHS Help with Health Costs services administered by NHSBSA are typically focused on support for English residents. Within this report figures represent all activity, with additional breakdowns by country available in the supporting datasets.

Integrated Care Board (ICB) areas have been used within this narrative to allow comparisons between areas. Applicants are aligned to an ICB area based on their postcode and mappings available using the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about).

------------------------------------------------------------------------

## 2. Results and commentary

### 2.1 Maternity exemption certificate

`r infoBox_border("",text = "Maternity exemption certifications exempt the holders from NHS prescription charges. It can also be used as proof of entitlement to free NHS dental treatment.

Maternity exemption certificates are available to anybody who is currently pregnant, or has given birth in the last 12 months. The certificate is valid until 12 months after the birth of the baby.

Applications are submitted via a midwife, doctor or health visitor who confirms eligibility. Both paper and digital applications are available.

As a maternity exemption certificate provides support with NHS prescription charges, support is generally applicable to applicants from England only." 
, width = "100%")`

#### Number of maternity exemption certificates issued 2018/19 to 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b> There were 552,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Issued certificates increased 12% from 2022/23 to 2023/24</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 1: The number of certificates issued has increased each year since 2020/21

Number of maternity exemption certificates issued from 2018/19 to 2023/24

```{r matex_issued_chart}
mat_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = mat_issued_objs$chart_data, filename = "hwhc_figure_1")
```

##### Table

###### Table 1: The number of certificates issued has increased each year since 2020/21

Number of maternity exemption certificates issued from 2018/19 to 2023/24

```{r matex_issued_table}
mat_issued_objs$table
get_download_button(title = "Download table data", data = mat_issued_objs$chart_data, filename = "hwhc_table_1")
```

</div>

Applications are submitted via a healthcare professional, who confirms eligibility. Therefore, only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

The number of issued maternity exemption certificates decreased in 2020/2021 but has increased in each of the following years. In 2023/24, the increase in certificates issued of 12% is the largest in recent years.

In 2023/24 there were 552,000 maternity exemption certificates issued, up from 494,000 in 2022/23.

#### Number of maternity exemption certificates issued, split by age band of recipient 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>Numbers peak in the 30-34 age band, with 198,000 certificates issued to this age group</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Distribution by age is similar to the pattern seen for live birth statistics</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 2: Certificates issued peaked in the 30-34 age group

Number of maternity exemption certificates issued in 2023/24 by age of recipient

```{r matex_age_chart}
mat_age_objs$chart
get_download_button(title = "Download chart data", data = mat_age_objs$chart_data, filename = "hwhc_figure_2")
```

##### Supporting Information (Live Births)
###### Figure 2 (supporting information): Live births peaked in the 30-34 age group

Number of live births in 2021, by age of mother

```{r matex_age_support_chart}
ch_births_age
get_download_button(title = "Download chart data", data = mat_age_objs$chart_data, filename = "hwhc_figure_2")
```

Source: Live birth figures from the Office for National Statistics

##### Table

###### Table 2: Certificates issued peaked in the 30-34 age group

Number of maternity exemption certificates issued in 2023/24 by age of recipient

```{r matex_age_table}
mat_age_objs$table
get_download_button(title = "Download table data", data = mat_age_objs$chart_data, filename = "hwhc_table_2")
```

</div>

The distribution of maternity exemption certificates by age peaks at 30-34, with 198,000 certificates issued to this age group in 2023/24. 

There is no publicly available data that accurately shows the number of women eligible for maternity exemptions in a specific time period. However, the age distribution is very similar to latest available live birth statistics published by Office for National Statistics.

#### Number of maternity exemption certificates issued, split by deprivation profile 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>More certificates were issued to people from the second most deprived quintile than any other</b>", width = "100%")`
:::
::: col-md-6
`r infoBox_no_border("", text = "<b>In the most deprived areas the number of certificates issued may be lower than expected compared to live birth figures alone</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 3: Certificates issued generally decrease in line with deprivation

Number of maternity exemption certificates issued in 2023/24 by IMD quintile of recipient

```{r mat_imd_chart}
mat_imd_objs$chart
get_download_button(title = "Download chart data", data = mat_imd_objs$chart_data, filename = "hwhc_figure_3")
```

##### Supporting Information (Live Births)
###### Figure 3 (supporting information): The number of live births decreases in line with deprivation

Number of live births in 2021, by IMD quintile of mother

```{r mat_imd_support_chart}
ch_births_imd
get_download_button(title = "Download chart data", data = mat_imd_objs$chart_data, filename = "hwhc_figure_3")
```

Source: Live birth figures from the Office for National Statistics

##### Table

###### Table 3: Certificates issued generally decrease in line with deprivation

Number of maternity exemption certificates issued in 2023/24 by IMD quintile of recipient

```{r mat_imd_table}
mat_imd_objs$table
get_download_button(title = "Download table data", data = mat_imd_objs$chart_data, filename = "hwhc_table_3")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

In general the number of maternity exemption certificates decreases in line with deprivation. The exception is the most deprived areas, where the 118,000 certificates issued is slightly less than 123,000 issued to the second most deprived group.

The latest available live birth statistics published by Office for National Statistics show that the number of births are higher in the more deprived areas. In the most deprived areas, women may be more likely to qualify for other income-related benefits that offer the same support as a maternity exemption certificate.

#### Number of maternity exemption certificates issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

##### Map

###### Figure 4 (map): Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2023/24, within ICB areas, per 10,000 female population aged 15 to 45

```{r mat_icb_map}
mat_icb_objs$map
get_download_button(title = "Download map data", data = mat_icb_objs$chart_data, filename = "hwhc_figure_4")
```

##### Chart

###### Figure 4 (chart): Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2023/24, within ICB areas, per 10,000 female population aged 15 to 45

```{r mat_icb_chart}
mat_icb_objs$chart
get_download_button(title = "Download chart data", data = mat_icb_objs$chart_data, filename = "hwhc_figure_4")
```

##### Table

###### Table 4: Certificates issued per 10,000 population varies by ICB area

Number of maternity exemption certificates issued in 2023/24, within ICB areas, per 10,000 female population aged 15 to 45

```{r mat_icb_table}
mat_icb_objs$table
get_download_button(title = "Download table data", data = mat_icb_objs$chart_data, filename = "hwhc_table_4")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. This is the latest available population statistics at the time of publication. To align with women who may benefit from maternity exemption certificates, population figures are based on the female population aged 15 to 45. However, these population statistics only cover a single year and will include women who are not pregnant and have not recently given birth.

Bedfordshire, Luton and Milton Keynes ICB had the highest estimated number of certificates issued per 10,000 population. In 2023/24, an estimated 637 certificates were issued per 10,000 population in this ICB.

Cheshire and Merseyside ICB had the lowest estimated number of certificates issued per 10,000 population in 2023/24. This ICB had an estimated 408 certificates per 10,000 population.

------------------------------------------------------------------------

### 2.2 Medical exemption certificate

`r infoBox_border("",text = "Medical exemption certifications exempt the holders from NHS prescription charges only. Patients with any of the qualifying conditions can apply for a certificate. But it is the certificate that provides exemption - not the condition.

Patients can apply for a medical exemption certificate if they have either:

* a permanent fistula (for example, caecostomy, colostomy, laryngostomy or ileostomy) which needs continuous surgical dressing or an appliance
* a form of hypoadrenalism (for example, Addison’s Disease) for which specific substitution therapy is essential
* diabetes insipidus and other forms of hypopituitarism
* diabetes mellitus, except where treatment is by diet alone
* hypoparathyroidism
* myasthenia gravis
* myxoedema (hypothyroidism which needs thyroid hormone replacement)
* epilepsy which needs continuous anticonvulsive therapy
* a continuing physical disability which means you cannot go out without the help of another person
* cancer and are undergoing treatment for either:
  * cancer
  * the effects of cancer
  * the effects of cancer treatment

A GP or doctor can provide the application form to qualifying patients. Medical exemption certificates are usually valid for 5 years. Certificates will not be automatically renewed upon expiration, with certificate holders needing to reapply.

As a medical exemption certificate provides support with NHS prescription charges, support is generally applicable to applicants from England only."
, width = "100%")`

#### Number of medical exemption certificates issued 2018/19 to 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>There were 475,000 certificates issued in 2023/24</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Certificates issued dropped during 2020/21 and 2021/22 but have since increased</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 5: The number of certificates issued has increased each year since 2020/21

Number of medical exemption certificates issued from 2018/19 to 2023/24

```{r medex_issued_chart}
med_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = med_issued_objs$chart_data, filename = "hwhc_figure_5")
```

##### Table

###### Table 5: The number of certificates issued has increased each year since 2020/21

Number of medical exemption certificates issued from 2018/19 to 2023/24

```{r medex_issued_table}
med_issued_objs$table
get_download_button(title = "Download table data", data = med_issued_objs$chart_data, filename = "hwhc_table_5")
```

</div>

Applications are authorised by a healthcare professional, who confirms eligibility. Therefore, only a small proportion of applications would not end with a certificate being issued. Certificates may not be issued if essential details cannot be confirmed by the applicant or other issues cannot be resolved. Due to the time taken to resolve an application, some certificates may be issued in a later time period to when the application was submitted.

The number of issued medical exemption certificates dipped in 2020/2021 but has since increased to the highest level in recent years.

In 2023/24 there were 475,000 medical exemption certificates issued, up from 453,000 in 2022/23.

#### Number of medical exemption certificates issued, split by age of recipient 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>134,000 certificates were issued to people aged 55-59 in 2023/24</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 6: The number of certificates issued rises as the age of recipient rises

Number of medical exemption certificates issued in 2023/24 by age of recipient

```{r medex_age_chart}
med_age_objs$chart
get_download_button(title = "Download chart data", data = med_age_objs$chart_data, filename = "hwhc_figure_6")
```

##### Table

###### Table 6: The number of certificates issued rises as the age of recipient rises

Number of medical exemption certificates issued in 2023/24 by age of recipient

```{r medex_age_table}
med_age_objs$table
get_download_button(title = "Download table data", data = med_age_objs$chart_data, filename = "hwhc_table_6")
```

</div>

There is a strong relationship between age and medical exemption certificates. The number of certificates issued steadily rises in line with age, which may be expected due to long-term conditions and multi-morbidity being more prevalent in older people. In 2023/24 the number of issued certificates ranged from 8,950 in the 15-19 age group, to 134,000 in the 55-59 age group.

As a medical exemption certificate only entitles the holder to free NHS prescriptions, it is not required after the age of 60, when NHS prescriptions are automatically free.

#### Number of medical exemption certificates issued, split by deprivation profile 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The number of medical exemption certificates issued in the most deprived quintile was lower than the next 2 quintiles with less deprivation</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 7: Certificates are issued less frequently to people in the least deprived areas

Number of medical exemption certificates issued in 2023/24 by IMD quintile

```{r medex_imd_chart}
med_imd_objs$chart
get_download_button(title = "Download chart data", data = med_imd_objs$chart_data, filename = "hwhc_figure_7")
```

##### Table

###### Table 7: Certificates are issued less frequently to people in the least deprived areas

Number of medical exemption certificates issued in 2023/24 by IMD quintile

```{r medex_imd_table}
med_imd_objs$table
get_download_button(title = "Download table data", data = med_imd_objs$chart_data, filename = "hwhc_table_7")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

Excluding the most deprived areas, as deprivation decreases so does the number of issued certificates.

The number of certificates issued in the most deprived areas was slightly lower than that for the next 2, less deprived, groups. In the most deprived areas people may be more likely to qualify for other income-related benefits that offer the same support as a medical exemption certificate.

#### Number of medical exemption certificates issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

##### Map

###### Figure 8 (map): Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_map}
med_icb_objs$map
get_download_button(title = "Download map data", data = med_icb_objs$chart_data, filename = "hwhc_figure_8")
```

##### Chart

###### Figure 8 (chart): Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_chart}
med_icb_objs$chart
get_download_button(title = "Download chart data", data = med_icb_objs$chart_data, filename = "hwhc_figure_8")
```

##### Table

###### Table 8: Certificates issued per 10,000 population varies by ICB area

Number of medical exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r med_icb_table}
med_icb_objs$table
get_download_button(title = "Download table data", data = med_icb_objs$chart_data, filename = "hwhc_table_8")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24, and will include patients who do not qualify for support via the medical exemption certificate. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to an ICB or age group from the available data.

There is not a large variation between ICBs. Values range from 220 certificates per 10,000 population in Greater Manchester ICB to 311 per 10,000 population in Lincolnshire ICB.

------------------------------------------------------------------------

### 2.3 NHS tax credit exemption certificates

`r infoBox_border("",text = "NHS tax credit exemption certificates provide support for:

* free NHS prescriptions
* free NHS dental treatment
* free NHS sight tests
* help with the cost of glasses or contact lenses
* free NHS wigs and fabric supports
* help with the cost of travel to receive NHS treatment

There is no application process as HM Revenue and Customs will identify who is eligible. Certificates are issued to people if their annual income used to calculate Tax Credits is £15,276 or less and they receive either:

* Child Tax Credit
* Working Tax Credit and Child Tax Credit paid together
* Working Tax Credit including a disability or severe disability element

Certificates are valid for 7 months, and issued every 6 months in April and September.

The majority of activity is for certificates issued to people in England, however certificates are also processed for people from Scotland, Wales and Northern Ireland."
, width = "100%")`

#### Number of NHS tax credit exemption certificates issued 2018/19 to 2023/24 {.tabset}

<div class = "tabset">

::: col-md-12
`r infoBox_no_border("", text = "<b>There were 1.2 million certificates issued in 2023/24</b>", width = "100%")`
:::

##### Chart

###### Figure 9: The number of certificates issued has been dropping year on year

Number of NHS tax credit exemption certificates issued from 2018/19 to 2023/24

```{r tax_issued_chart}
tax_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = tax_issued_objs$chart_data, filename = "hwhc_figure_9")
```

##### Table

###### Table 9: The number of certificates issued has been dropping year on year

Number of NHS tax credit exemption certificates issued by financial year from 2018/19 to 2023/24

```{r tax_issued_table}
tax_issued_objs$table
get_download_button(title = "Download table data", data = tax_issued_objs$chart_data, filename = "hwhc_table_9")
```

</div>

The number of issued Tax Credit Exemption Certificates has been dropping year on year from 3.7 million in 2018/19 to 1.2 million in 2023/24. This strong decline may be linked to the Universal Credit roll-out as people are migrated to new benefit schemes. The number of certificates issued each year has declined by between 22% and 26% since 2019/2020.

#### Number of NHS tax credit exemption certificates issued, split by age band 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>77% of certificates were issued to people aged 35 to 54</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 10: The number of certificates issued peaks at age 40-44

Number of NHS tax credit exemption certificates issued in 2023/24, split by age of recipient

```{r tax_age_chart}
tax_age_objs$chart
get_download_button(title = "Download chart data", data = tax_age_objs$chart_data, filename = "hwhc_figure_10")
```

##### Table

###### Table 10: The number of certificates issued peaks at age 40-44

Number of NHS tax credit exemption certificates issued in 2023/24, split by age of recipient

```{r tax_age_table}
tax_age_objs$table
get_download_button(title = "Download table data", data = tax_age_objs$chart_data, filename = "hwhc_table_10")
```

</div>

The number of Tax Credit Exemption Certificates issued in 2023/24 peaked at ages 40-44 with 268,000 certificates issued.

Over three-quarters of all certificates were issued to people aged between 35 and 54.

Although numbers are relatively small, some certificates are issued to people aged 60 and over as Tax Credit Exemption Certificate also provides support with some health costs which do not have an age exemption.

#### Change in age profile of NHS tax credit exemption certificates issued 2018/19 to 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2018/19, over 55% of NHS tax credit exemption certificates were issued to people under 40 years old</b>", width = "100%")`
:::
::: col-md-6
`r infoBox_no_border("", text = "<b>By 2023/24 this had dropped to just over 30% of issued certificates</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 11: People aged under 40 are representing less of the NHS tax credit exemption population each year

Number of NHS tax credit exemption certificates issued by financial year, 2018/19 to 2023/24, split by age group of recipient

```{r tax_age_trend_chart}
ch_tax_age_trend
get_download_button(title = "Download chart data", data = dl_tax_age_trend, filename = "hwhc_figure_11")
```

##### Table

###### Table 11: People aged under 40 are representing less of the NHS tax credit exemption population each year

Number of NHS tax credit exemption certificates issued by financial year, 2018/19 to 2023/24, split by age group of recipient

```{r tax_age_trend_table}
tbl_tax_age_trend
get_download_button(title = "Download table data", data = dl_tax_age_trend, filename = "hwhc_table_11")
```

</div>

The age profile of Tax Credit Exemption Certificates have changed considerably over the past few years. In 2018/19 over 55% of certificates holders were under 40 years old, but as of 2023/24 this group represents only 30% of certificates issued. Following the introduction of Universal Credit, new applicants will receive support via Universal Credit rather than tax credits.

#### Number of NHS tax credit exemption certificates issued by deprivation profile 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The number of certificates issued to the most deprived areas is almost 6 times the number issued in the least deprived areas</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 12: Certificates are most commonly issued to the most deprived areas

Number of NHS tax credit exemption certificates issued in 2023/24, split by IMD quintile

```{r tax_imd_chart}
tax_imd_objs$chart
get_download_button(title = "Download chart data", data = tax_imd_objs$chart_data, filename = "hwhc_figure_12")
```

##### Table

###### Table 12: Certificates are most commonly issued to the most deprived areas

Number of NHS tax credit exemption certificates issued in 2023/24, split by IMD quintile

```{r tax_imd_table}
tax_imd_objs$table
get_download_button(title = "Download table data", data = tax_imd_objs$chart_data, filename = "hwhc_table_12")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure. These indices of deprivation are specific to England and therefore can only be applied to certificates that can be assigned to an English postcode.

There is a strong relationship between deprivation and number of NHS tax credit exemption certificates issued. This is to be expected as this is a means-tested benefit. In 2023/24, there were 394,000 certificates issued in the most deprived areas. In the least deprived areas only 67,800 certificates were issued in 2023/24.

#### Number of NHS tax credit exemption certificates issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

##### Map

###### Figure 13 (map): Certificates issued per 10,000 population varies by ICB area

Number of NHS tax credit exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r tax_icb_map}
tax_icb_objs$map
get_download_button(title = "Download map data", data = tax_icb_objs$chart_data, filename = "hwhc_figure_13")
```

##### Chart

###### Figure 13 (chart): Certificates issued per 10,000 population varies by ICB area

Number of NHS tax credit exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r tax_icb_chart}
tax_icb_objs$chart
get_download_button(title = "Download chart data", data = tax_icb_objs$chart_data, filename = "hwhc_figure_13")
```

##### Table

###### Table 13: Certificates issued per 10,000 population varies by ICB area

Number of NHS tax credit exemption certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r tax_icb_table}
tax_icb_objs$table
get_download_button(title = "Download table data", data = tax_icb_objs$chart_data, filename = "hwhc_table_13")
```

</div>

ICB areas have been used to allow comparisons between areas, however ICBs are not responsible for NHS tax credit exemption certificates. Applicants have been associated with an ICB area based on their postcode, which is only possible for people within England.

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. This is the latest available population statistics at the time of publication. To align with NHS tax credit exemption certificates, population figures are based on the population aged 16 and over.

There are a small number of ICBs where the number of certificates per 10,000 population is much higher than the rest of the country.

In 2023/24, North East London ICB had 510 certificates issued per 10,000 population. This was closely followed by 507 certificates per 10,000 population in Birmingham and Solihull ICB. Numbers then drop to 386 certificates per 10,000 population in North Central London ICB.

In 2023/24, Bath and North East Somerset, Swindon and Wiltshire ICB had the lowest rate, with only 85 certificates issued per 10,000 population.

------------------------------------------------------------------------

### 2.4 NHS Low Income Scheme

`r infoBox_border("",text = "The NHS Low Income Scheme (LIS) provides income related help. Support is available for people who do not receive help with health costs via an existing exemption. This is a means tested benefit, but also considers council tax and housing costs.

NHS Low Income Scheme exemption provides help with:

* NHS prescriptions
* NHS dental treatment
* sight tests, glasses and contact lenses
* travel to receive NHS treatment
* NHS wigs and fabric supports

The amount of support provided is based on weekly income, necessary outgoings, plus any savings or investments. Issued certificates may also cover the applicant's partner and dependent children under 20 years old.

The majority of certificates are issued for a year, but some are issued for periods of between 1 day and 5 years, depending on circumstances.

The majority of activity is for applicants from England, however applications are also processed for applicants from Scotland and Wales.

These statistics exclude any figures for certificates issued to asylum seekers assisted by UK Visas and Immigration (UKVI). These individuals are issued certificates automatically as part of the support they receive and are not handled through the regular application and assessment process."
, width = "100%")`

#### Number of NHS Low Income Scheme certificates issued by certificate type 2018/19 to 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>The number of certificates issued decreased in 2023/24</b>", width = "100%")`
:::
:::

Applicants will either complete a paper HC1 form or apply via the [online application for LIS](https://apply-for-help-with-nhs-costs.nhsbsa.nhs.uk/apply-online). Two levels of support are available from the NHS Low Income Scheme. Depending on the outcome of the assessment, applicants could receive:

* an [HC2 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc2-certificates-full-help-health-costs) for full help with health costs, including free NHS prescriptions
* an [HC3 certificate](https://www.nhsbsa.nhs.uk/nhs-low-income-scheme/hc3-certificates-limited-help-health-costs) for limited help with health costs, which will show how much the holder has to pay towards health costs

##### Chart

###### Figure 14: Almost two-thirds of certificates issued provided full help with health costs

Number of NHS Low Income Scheme certificates issued, by certificate type, from 2018/19 to 2023/24

```{r lis_issued_chart}
lis_issued_objs$chart |> 
  highcharter::hc_yAxis(min = 0)
get_download_button(title = "Download chart data", data = lis_issued_objs$chart_data, filename = "hwhc_figure_14")
```

##### Table

###### Table 14: Almost two-thirds of certificates issued provided full help with health costs

Number of NHS Low Income Scheme certificates issued, by certificate type, from 2018/19 to 2023/24

```{r lis_issued_table}
lis_issued_objs$table
get_download_button(title = "Download table data", data = lis_issued_objs$chart_data, filename = "hwhc_table_14")
```

</div>

There were 249,000 certificates issued in 2023/24. This included 161,000 HC2 certificates and 87,100 HC3 certificates. Compared to 2022/23 there has been a slight decrease in the number of certificates issued to customers. 

A certificate is only issued to the applicant following the completion of the assessment. Therefore these figures do not include any applications that are ongoing or were abandoned before completion. These figures also exclude any completed applications where no certificate was issued to the applicant.

#### Number of NHS Low Income Scheme certificates issued, split by certificate type and applicant age band 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24 more certificates were issued to applicants aged 65 and over than any other age group</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 15: Over half of certificates issued were to people aged under 25 or 65 and over

Number of NHS Low Income Scheme certificates issued in 2023/24, by certificate type and applicant age band

```{r lis_age_chart}
lis_age_objs$chart
get_download_button(title = "Download chart data", data = lis_age_objs$chart_data, filename = "hwhc_figure_15")
```

##### Table

###### Table 15: Over half of certificates issued were to people aged under 25 or 65 and over

Number of NHS Low Income Scheme certificates issued in 2023/24, by certificate type and applicant age band

```{r lis_age_table}
lis_age_objs$table
get_download_button(title = "Download table data", data = lis_age_objs$chart_data, filename = "hwhc_table_15")
```

</div>

NHS Low Income Scheme certificates cover the applicant, partners and any dependent children. Age information is only captured for the lead applicant though. 

Pensioners are more likely than other age groups to receive a long term certificate. This will impact the age distribution of certificates issued in a year as pensioners may only have to apply once every 5 years.

The younger and older age groups receive more certificates than other ages. Over 52,000 certificates went to people aged 65 and over in 2023/24. A further 51,000 went to applicants 20-24 and 32,000 to applicants aged 19 and under.

#### Number of NHS Low Income Scheme certificates issued, split by certificate type and IMD quintile 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24 70,000 certificates went to people in the most deprived areas</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>In 2023/24, the second quintile received almost as many HC2 certificates as the most deprived quintile</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 16: Certificates were issued three times more often in the most deprived areas than the least deprived areas

Number of NHS Low Income Scheme certificates issued in 2023/24, split by certificate type and IMD quintile

```{r lis_imd_chart}
lis_imd_objs$chart
get_download_button(title = "Download chart data", data = lis_imd_objs$chart_data, filename = "hwhc_figure_16")
```

##### Table

###### Table 16: Certificates were issued three times more often in the most deprived areas than the least deprived areas

Number of NHS Low Income Scheme certificates issued in 2023/24, split by certificate type and IMD quintile

```{r lis_imd_table}
lis_imd_objs$table
get_download_button(title = "Download table data", data = lis_imd_objs$chart_data, filename = "hwhc_table_16")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure. These indices of deprivation are specific to England and therefore can only be applied to certificates that can be assigned to an English postcode.

Certificates were issued in the most deprived areas more than 3 times more often than the least deprived areas. In the most deprived areas, people may be more likely to qualify for other income-related benefits that offer some of the same support as available through the NHS Low Income Scheme.

Despite being means-tested, certificates are still issued to applicants from less deprived areas. In 2023/24, over 13,000 HC2 certificates and over 8,000 HC3 certificates went to the least deprived areas.

#### Number of NHS Low Income Scheme certificates issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>North London had 3 of the top 4 ICBs in terms of certificates issued per 10,000 population</b>", width = "100%")`
:::
:::

The map, chart and table are based on the total combined number of HC2 and HC3 certificates issued in 2023/24. Figures for individual certificate types are included in the accompanying data.

##### Map

###### Figure 17 (map): Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r lis_icb_map}
lis_icb_objs$map
get_download_button(title = "Download map data", data = lis_icb_objs$chart_data, filename = "hwhc_figure_17")
```

##### Chart

###### Figure 17 (chart): Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r lis_icb_chart}
lis_icb_objs$chart
get_download_button(title = "Download chart data", data = lis_icb_objs$chart_data, filename = "hwhc_figure_17")
```

##### Table

###### Table 17: Certificates issued per 10,000 population varies by ICB area

Number of NHS Low Income Scheme certificates issued in 2023/24, within ICB areas, per 10,000 population aged 16 and over

```{r lis_icb_table}
lis_icb_objs$table
get_download_button(title = "Download table data", data = lis_icb_objs$chart_data, filename = "hwhc_table_17")
```

</div>

ICB areas have been used to allow comparisons between areas, however ICBs are not responsible for applications to the NHS Low Income Scheme. Applicants have been aligned to an ICB area based on their postcode, which is only possible for applicants within England.

Population estimates use figures published by [Office for National Statistics for ICB population](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). These figures are mid-year estimates with 2022 population estimates used for this report. These are the latest available population statistics at the time of publication. To align with the people who may benefit from the NHS Low Income Scheme, population figures are based on people aged 16 and over.

North West London ICB had the highest estimated number of certificates issued per 10,000 population. In 2023/24, an estimated 85 certificates were issued per 10,000 population in this ICB. North London had 3 of the top 4 ICBs in terms of certificates issued per 10,000 population.

Bath and North East Somerset, Swindon and Wiltshire ICB had the lowest estimated number of certificates issued per 10,000 population in 2023/24. This ICB had an estimated 29 certificates per 10,000 population.

------------------------------------------------------------------------

### 2.5 NHS Prescription Prepayment Certificate (PPC)

`r infoBox_border("",text = "A PPC can help provide savings for people who pay for NHS Prescriptions. The certificate covers all a patient's NHS prescription charges for a set price. Patients will save money if they need 4 or more items in 3 months or 12 or more items in 12 months.

PPCs can be purchased in-person at some pharmacies or direct via the [PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-prescription-prepayment-certificate-ppc). Applicants can choose between a 3-month and 12-month certificate.

Patients receiving prescriptions for hormone replacement therapy (HRT) may benefit more from the Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC).

As a PPC provides support with NHS prescription charges, support is generally applicable to applicants from England only."
, width = "100%")`

#### Number of PPCs issued 2018/19 to 2023/24, split by PPC type {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>There were 3.0 million certificates issued in 2023/24</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 18: The number of PPCs issued has increased each year

Number of PPCs issued from 2018/19 to 2023/24, split by PPC type

```{r ppc_issued_chart}
ppc_issued_objs$chart
get_download_button(title = "Download chart data", data = ppc_issued_objs$chart_data, filename = "hwhc_figure_18")
```

##### Table

###### Table 18: The number of PPCs issued has increased each year

Number of PPCs issued from 2018/19 to 2023/24, split by PPC type

```{r ppc_issued_table}
ppc_issued_objs$table
get_download_button(title = "Download table data", data = ppc_issued_objs$chart_data, filename = "hwhc_table_18")
```

</div>

Not all applications will result in a certificate being issued. Certificates will not be issued if there are issues with the initial payment for the certificate and in some cases the applicant may drop out for other reasons.

The number of 12-month PPCs issued has shown a consistent year-on-year increase. In 2023/24 there were 2.1 million certificates issued, up from 2.0 million in 2022/23 which is a 4% increase. The number of 12-month certificates issued is more than double the number of 3-month certificates issued. 

The year-on-year increase for 3-month certificates issued is not as high, and the number of certificates issued dropped in 2020/21. In 2023/24 there were 936,000 certificates issued, up 1% from the 923,000 certificates issued in 2022/23.

#### Number of PPCs issued by age of recipient, split by PPC type 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24, there were 782,000 certificates issued to people aged 55-59</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 19: The number of certificates issued rises as the age of the recipient rises

Number of PPCs issued in 2023/24, split by PPC type and age of recipient

```{r ppc_age_chart}
ppc_age_objs$chart
get_download_button(title = "Download chart data", data = ppc_age_objs$chart_data, filename = "hwhc_figure_19")
```

##### Table

###### Table 19: The number of certificates issued rises as the age of the recipient rises

Number of PPCs issued in 2023/24, split by PPC type and age of recipient

```{r ppc_age_table}
ppc_age_objs$table
get_download_button(title = "Download table data", data = ppc_age_objs$chart_data, filename = "hwhc_table_19")
```

</div>

There is a strong relationship between age and number of PPCs issued. 

In 2023/24, the number of 12-month certificates issued ranged from 19,900 for people 15-19 to 527,000 for people aged 55-59. There is a similar pattern for 3-month certificates but numbers are lower, ranging from 13,500 in the 15-19 age group to 255,000 for people aged 55-59.

As a PPC only provides support for NHS prescription charges, it is not required after the age of 60, when NHS prescriptions are automatically free. For applicants who choose to auto-renew their certificate, the certificate will not be renewed for people turning 60 within 9 months of their existing certificate expiring.

The 12-month certificate is more popular across all age bands. However, the ratio of 3-month to 12-month certificates generally increases in line with age. In the 15-19 age group the number of 12-month PPCs is only 1.5 times the number of 3-month PPCs. In the 50-54 age group this rate rises to 2.8 before dropping back down to 2.0 for the 55-59 age group.

#### Number of PPCs issued by deprivation profile, split by PPC type 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
:::

##### Chart

###### Figure 20: Certificates are issued less frequently in the most and least deprived areas

Number of PPCs issued in 2023/24, split by PPC type and IMD quintile

```{r ppc_imd_chart}
ppc_imd_objs$chart
get_download_button(title = "Download chart data", data = ppc_imd_objs$chart_data, filename = "hwhc_figure_20")
```

##### Table

###### Table 20: Certificates are issued less frequently in the most and least deprived areas

Number of PPCs issued in 2023/24, split by PPC type and IMD quintile

```{r ppc_imd_table}
ppc_imd_objs$table
get_download_button(title = "Download table data", data = ppc_imd_objs$chart_data, filename = "hwhc_table_20")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

There is not a strong relationship between deprivation and number of PPCs issued. However, numbers are lower in the most and least deprived areas of the country. In the most deprived areas, people may be more likely to qualify for other income-related benefits that provide access to free NHS prescriptions.

#### Number of PPCs issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>In 2023/24 the 5 London ICBs were the bottom 5 in terms of certificates issued per 10,000 population</b>", width = "100%")`
:::
:::

The map, chart and table are based on the total number of 3-month and 12-month certificates issued in 2023/24. Figures for individual certificate types are included in the accompanying data.

##### Map

###### Figure 21 (map): Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_map}
ppc_icb_objs$map
get_download_button(title = "Download map data", data = ppc_icb_objs$chart_data, filename = "hwhc_figure_21")
```

##### Chart

###### Figure 21 (chart): Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_chart}
ppc_icb_objs$chart
get_download_button(title = "Download chart data", data = ppc_icb_objs$chart_data, filename = "hwhc_figure_21")
```

##### Table

###### Table 21: Certificates issued per 10,000 population varies by ICB area

Number of PPCs issued in 2023/24, within ICB areas, per 10,000 population aged 16 to 59 receiving NHS prescriptions

```{r ppc_icb_table}
ppc_icb_objs$table
get_download_button(title = "Download table data", data = ppc_icb_objs$chart_data, filename = "hwhc_table_21")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to a ICB or age group from the available data.

In 2023/24 the 5 London ICBs were the bottom 5 in terms of certificates issued per 10000 population. The lowest figure was for North Central London ICB, with 830 certificates issued per 10,000 population in 2023/24. 

Lincolnshire ICB had the highest figure in 2023/24, with 2,659 certificates issued per 10,000 population.

------------------------------------------------------------------------

### 2.6 NHS Hormone Replacement Therapy Prescription Prepayment Certificate (HRT PPC)

`r infoBox_border("",text = "The HRT PPC was introduced in April 2023 to provide support with the prescription costs for medication used to treat symptoms of the menopause. 

A HRT PPC covers all a patient's NHS prescription charges, for certain HRT medicines, for a set price. HRT PPCs are valid for 12 months and patients will save money if they need 3 or more HRT prescription items in 12 months.

The HRT PCC can be used for any of the applicable HRT medicines, regardless of why they are prescribed. The HRT PPC webpages include a [list of the medicines covered by a HRT PPC](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc/medicines-covered-hrt-ppc).

HRT PPCs can be purchased in-person at some pharmacies or direct via the [HRT PPC webpage](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc).

Patients receiving prescriptions for other medicines, including HRT medicines not covered by the HRT PPC, may benefit more from a 3 or 12 month Prescription Prepayment Certificate (PPC) that covers all NHS prescriptions.

As a HRT PPC provides support with NHS prescription charges, support is generally applicable to applicants from England only."
, width = "100%")`

#### Number of HRT PPCs issued 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>In the first year it was available, 566,000 HRT PPCs were issued</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>Almost half of the certificates issued in March 2024 were set to start in April 2024</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 22: Certificates issued dropped following a strong launch in April 2023 but have started to climb

Number of HRT PPCs issued by month during 2023/24

```{r hrt_issued_chart}
hrt_issued_objs$chart
get_download_button(title = "Download chart data", data = hrt_issued_objs$chart_data, filename = "hwhc_figure_22")
```

##### Supporting Information

###### Figure 22 (post-dated certicates): March 2024 had a spike in the proportion of certificates post-dated to start the following month

Proportion of issued HRT PPCs post-dated to start the following month, by month of certificate issue

```{r hrt_issued_support}
ch_hrt_issued_postdate
get_download_button(title = "Download chart data", data = hrt_issued_objs$chart_data, filename = "hwhc_figure_22")
```

##### Table

###### Table 22: Certificates issued dropped following a strong launch in April 2023 but have started to climb

Number of HRT PPCs issued by month during 2023/24

```{r hrt_issued_table}
hrt_issued_objs$table
get_download_button(title = "Download table data", data = hrt_issued_objs$chart_data, filename = "hwhc_table_22")
```

</div>

There were 566,000 HRT PPCs issued in 2023/24. Although numbers decreased after a very strong launch of the service, numbers have started to increase near the end of the year when certificates issued early in 2023/24 will be nearing expiration.

The 131,000 certificates issued in the first month represent 23% of all HRT PPCs issued in 2023/24.

The number of certificates issued leveled out near the end of the financial year with a notable jump in March 2024 when 37,600 certificates were issued. However, almost half of the certificates issued in March 2024 had a certificate start date in April 2024 which may suggest a large number of these were customers renewing expiring certificates. 

#### Number of HRT PPCs issued by age of recipient 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-6
`r infoBox_no_border("", text = "<b>92% of HRT PPCs were issued to recipients aged 45 and over</b>", width = "100%")`
:::

::: col-md-6
`r infoBox_no_border("", text = "<b>The age group with the most HRT PPCs issued was 50-54</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 23: Most people issued HRT PPCs were aged 45 and over

Number of HRT PPCs issued in 2023/24, split by age of recipient


```{r hrt_age_chart}
hrt_age_objs$chart
get_download_button(title = "Download chart data", data = hrt_age_objs$chart_data, filename = "hwhc_figure_23")
```

##### Supporting Information

###### Figure 23 (NHS prescribing): Most patients receiving HRT PPC eligible medication were aged 45 and over

Estimated number of patients receiving NHS prescribing in 2023/24 of medication qualifying for support by HRT PPC, split by age of recipient

```{r hrt_age_support}
hrt_age_base_pop_objs$chart
get_download_button(title = "Download chart data", data = hrt_age_objs$chart_data, filename = "hwhc_figure_23")
```

##### Table

###### Table 23: Most people issued HRT PPCs were aged 45 and over

Number of HRT PPCs issued in 2023/24, split by age of recipient

```{r hrt_age_table}
hrt_age_objs$table
get_download_button(title = "Download table data", data = hrt_age_objs$chart_data, filename = "hwhc_table_23")
```

</div>

Around 92% of HRT PPCs were issued to people aged 45 to 59. This may be expected as the menopause usually affects women between the ages of 45 and 55. Certificates peaked at the 50 to 54 age group, with 229,000 HRT PPCs issued to this age group.

The distribution of HRT PPCs by age is similar to the distribution of the estimated number of patients receiving the HRT qualifying medication on NHS prescriptions in 2023/24.

#### Number of HRT PPCs issued by deprivation profile 2023/24 {.tabset}

<div class = "tabset">

::: {.row style="display: flex; padding-bottom: 15px;"}
::: col-md-12
`r infoBox_no_border("", text = "<b>Only 37,600 certificates were issued in the most deprived areas in 2023/24</b>", width = "100%")`
:::
:::

##### Chart

###### Figure 24: The number of HRT PPCs issued rises as deprivation decreases

Number of HRT PPCs issued in 2023/24, by IMD quintile

```{r hrt_imd_chart}
hrt_imd_objs$chart
get_download_button(title = "Download chart data", data = hrt_imd_objs$chart_data, filename = "hwhc_figure_24")
```

##### Supporting Information

###### Figure 24 (eligible patients): Patients in the least deprived areas are more likely to receiving prescribing of HRT medication

Estimated number of patients receiving NHS prescribing in 2023/24 of medication qualifying for support by HRT PPC, split by IMD quintile

```{r hrt_imd_support}
hrt_imd_base_pop_objs$chart
get_download_button(title = "Download chart data", data = hrt_imd_objs$chart_data, filename = "hwhc_figure_24")
```

##### Table

###### Table 24: The number of HRT PPCs issued rises as deprivation decreases

Number of HRT PPCs issued in 2023/24, by IMD quintile

```{r hrt_imd_table}
hrt_imd_objs$table
get_download_button(title = "Download table data", data = hrt_imd_objs$chart_data, filename = "hwhc_table_24")
```

</div>

The English Indices of Deprivation provide a measure of deprivation. In this report deprivation reflects the applicant's registered address. [Section 3.3 of this summary](#about_imd) contains more information on this measure.

There is a strong relationship between deprivation and the number of HRT PPCs issued in 2023/24. There were only 37,600 certificates issued in the most deprived areas. There were 189,000 certificates issued in the least deprived areas.

The distribution of HRT PPCs by deprivation is similar to the distribution of the estimated number of patients receiving the HRT qualifying medication on NHS prescriptions in 2023/24. However, the relationship between deprivation and issued certificates is a bit more pronounced for HRT PPCs issued, with less certificates issued to the most deprived areas. In the most deprived areas, people may be more likely to qualify for other methods of help with health costs.

#### Number of HRT PPCs issued per 10,000 population, by ICB 2023/24 {.tabset}

<div class = "tabset">

##### Map

###### Figure 25 (map): A cluster of ICBs in the South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2023/24 within ICB areas per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_map}
hrt_icb_objs$map
get_download_button(title = "Download map data", data = hrt_icb_objs$chart_data, filename = "hwhc_figure_25")
```

##### Chart

###### Figure 25 (chart): A cluster of ICBs in the South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2023/24 within ICB areas per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_chart}
hrt_icb_objs$chart
get_download_button(title = "Download chart data", data = hrt_icb_objs$chart_data, filename = "hwhc_figure_25")
```

##### Table

###### Table 25: A cluster of ICBs in the South West have higher rates than the rest of the country

Number of HRT PPCs issued in 2023/24 within ICB areas per 10,000 patients receiving NHS prescribing of eligible medication

```{r hrt_icb_table}
hrt_icb_objs$table
get_download_button(title = "Download table data", data = hrt_icb_objs$chart_data, filename = "hwhc_table_25")
```

</div>

ICB areas have been used to allow comparisons between areas, with applicants aligned to an ICB area based on their postcode.

Population estimates are based on the number of patients, aged 16-59, identified receiving NHS prescriptions for drugs covered by the HRT PPC during 2023/24. The population estimates will include all patients regardless of whether or not they paid prescription charges in 2023/24. These population estimates may miss some patients who cannot be identified in the NHS prescriptions dataset, or those who cannot be assigned to a ICB or age group from the available data.

Values range from 2,700 certificates per 10,000 population in Black Country ICB to 4,034 per 10,000 population in South West London ICB.

In 2023/24, a cluster of ICBs in the South West region had a higher number of HRT PPCs issued per 10,000 population than other ICBs.

------------------------------------------------------------------------

## 3 About these statistics

### 3.1 Statistical Disclosure Control (SDC) {.toc-ignore}

Statistical disclosure control has been applied to these statistics. Volume figures have been redacted in the supporting summary tables if they relate to fewer than 5 items. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

### 3.2. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases rounding has been to 2 significant figures in the key findings and information boxes and 3 significant figures in the main text. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 3.3. Index of deprivation {#about_imd .toc-ignore}

The [English Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) have been used to provide a measure of deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals' living conditions.

The reported IMD quintile, where 1 is the most deprived and 5 the least deprived, is derived from the postcode currently held for an applicant. There are a small number of records each year that we have reported as having an unknown IMD quintile. These are applicants with no fixed abode or with british forces addresses, or where we have been unable to match the applicant postcode to an English postcode in the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about).

IMD quintiles are calculated by ranking census lower-layer super output areas (LSOAs) from most deprived to least deprived and dividing them into equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

### 3.4. ICB classification {#about_icb .toc-ignore}

ICB areas have been used to allow comparisons between areas. Applicants have been associated with an ICB area based on their postcode using the [National Statistics Postcode Lookup (NSPL) - May 2024](https://geoportal.statistics.gov.uk/datasets/521edce4159a451a932539b7fc786322/about). As ICBs are only available within England this geography classificaiton is only available for applicants with a valid English postcode.

------------------------------------------------------------------------

## 4 Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:
**Email**:
[accessibility\@nhsbsa.nhs.uk](mailto:accessibility@nhsbsa.nhs.uk){.email}

**Phone**: 0191 203 5318

[Find out about call
charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We'll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the 'Contact us' section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

------------------------------------------------------------------------

## 5 Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:**
[statistics\@nhsbsa.nhs.uk](mailto:statistics@nhsbsa.nhs.uk){.email}

**You can also write to us at:**

NHSBSA - Statistics\
NHS Business Services Authority\
Stella House\
Goldcrest Way\
Newburn Riverside\
Newcastle upon Tyne\
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>
